---
title: "aDNA damage analysis - full data range"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
# 1. Preparations
# 1.1 Packages
```{r}
library(tidyverse)
packageVersion('tidyverse')
library(patchwork)
packageVersion('patchwork')
library(ggrepel)
packageVersion('ggrepel')
library(propagate)
packageVersion('propagate')
library(pROC)
packageVersion('pROC')
library(scales)
packageVersion('scales')
library(nlstools)
```
# 1.2 Read the data

I worked with a training data set (data) to explore trends and fit models and a test dataset (test_genera) to test the models' performance

#### Training data
```{r}
data=read.csv('damage_data_core_genera.csv')
```
```{r}
data_w = pivot_wider(data[,-c(3:4)],
           names_from = tool_param,
           values_from = value )
data_w$mapdamage_1_over_lambda=1/data_w$mapdamage_lambda
data_w$pydamage_1over_p=1/data_w$pydamage_p
data_w$qPCRcopies_per_uL_and_mg_rock=data_w$qPCRcopies_per_uL/data_w$mass_mg
```
#### Test data
```{r}
test_genera = read_csv("damage_data_test_genera.csv")
```
```{r}
test_genera_w = pivot_wider(test_genera[,-c(3:4)],
           names_from = tool_param,
           values_from = value )
test_genera_w$qPCRcopies_per_uL_and_mg_rock=test_genera_w$qPCRcopies_per_uL/test_genera_w$mass_mg
```

# 2. Plot functions
```{r}
plot_saturation_curves=function(table, param){
  ggplot(table, aes(x = median_age_yrs_bf_1950)) +
    geom_ribbon(aes(x = median_age_yrs_bf_1950, ymin = CI_lower_full, ymax = CI_upper_full), alpha=0.1, fill = "darkred") +
    geom_ribbon(aes(x = median_age_yrs_bf_1950, ymin = CI_lower_trunc, ymax = CI_upper_trunc), alpha=0.1, fill = "black") +
  geom_line(aes(y= CI_lower_full), color="darkred", linetype='dotted')+
  geom_line(aes(y= CI_upper_full), color="darkred", linetype='dotted')+
   geom_line(aes(y= CI_lower_trunc), color="black", linetype='dotted')+
  geom_line(aes(y= CI_upper_trunc), color="black", linetype='dotted')+
  geom_point(aes(y = {{param}}), shape = 4, color='gray60', size=1) +
  geom_line(aes(y=fitted_full), color='darkred',linetype='solid')+
  geom_vline(xintercept = 32000, color='gray40')+
  geom_line( aes(y = fitted_trunc), color = "black",linetype='solid')+
  theme_bw(base_size=9)+
  scale_x_continuous(expand = c(0,0), limits=c(0,38000))+
  labs(x = "Nominal Age (Years BP)")
}
```

```{r}
plot_residuals=function(table){
ggplot(table)+
geom_point(aes(x=median_age_yrs_bf_1950, y=residual_full),shape=1,color="darkred", size=2)+
geom_point(aes(x=median_age_yrs_bf_1950, y=residual_trunc),shape=16, color="black", size=1)+
geom_hline(yintercept=0)+
geom_vline(xintercept=32000,color='gray40')+
geom_segment(aes(x = median_age_yrs_bf_1950, y = residual_full, xend = median_age_yrs_bf_1950, yend = residual_trunc))+
theme_bw(base_size = 9)+
labs(x='Nominal Age (Years BP)', y='Residual')
}

```



# 3. Exponential saturation models: Damage over time
## 3.1 delta_s over time
### 3.1.1 Fit
```{r}
plottable_delta_s=data_w[,c('median_age_yrs_bf_1950', 'mapdamage_delta_s', 'genus','sample','gc_gc')]
```
```{r}
fit_delta_s_trunc <- nls(mapdamage_delta_s  ~ 1 - exp(-r * median_age_yrs_bf_1950),data = plottable_delta_s[plottable_delta_s$median_age_yrs_bf_1950<32000,],start = list(r = 1e-3))
summary(fit_delta_s_trunc)
```

```{r}
fit_delta_s_full <- nls(mapdamage_delta_s  ~ 1 - exp(-r * median_age_yrs_bf_1950),data = plottable_delta_s,start = list(r = 1e-3))
 summary(fit_delta_s_full)
```

### 3.1.2 Confidence and Prediction interval
```{r,message=FALSE}
# Prediction intervals
#full data
CI_delta_s_full <- predictNLS(fit_delta_s_full,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_delta_s$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
#truncated data
CI_delta_s_trunc <- predictNLS(fit_delta_s_trunc,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_delta_s$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
```

### 3.1.3 Plot
```{r}
plottable_delta_s$CI_lower_full <- CI_delta_s_full[, "Sim.2.5%"]
plottable_delta_s$CI_upper_full <- CI_delta_s_full[, "Sim.97.5%"]
plottable_delta_s$fitted_full <- predict(fit_delta_s_full)
plottable_delta_s$residual_full=residuals(fit_delta_s_full)

plottable_delta_s$CI_lower_trunc <- CI_delta_s_trunc[, "Sim.2.5%"]
plottable_delta_s$CI_upper_trunc <- CI_delta_s_trunc[, "Sim.97.5%"]
plottable_delta_s$fitted_trunc = predict(fit_delta_s_trunc, newdata = plottable_delta_s)
plottable_delta_s$residual_trunc= plottable_delta_s$mapdamage_delta_s - predict(fit_delta_s_trunc, newdata = plottable_delta_s)

```
```{r}
p_deltas=plot_saturation_curves(plottable_delta_s, mapdamage_delta_s)
p_deltas
```


### 3.1.5 Residuals
```{r}
delta_s_res=plot_residuals(plottable_delta_s)

delta_s_res
```

  
## 3.2 pmax over time
### 3.2.1 Fit
```{r}
plottable_pmax=data_w[,c('median_age_yrs_bf_1950', 'pydamage_pmax','genus', 'sample', 'gc_gc')]

```
```{r}
# Fit the model using nls
fit_pmax_full <- nls(pydamage_pmax  ~ a+b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_pmax,
start = list(r = 1e-4, b=0.6,a=0.2))
summary(fit_pmax_full)
```

```{r}
plottable_pmax=data_w[,c('median_age_yrs_bf_1950', 'pydamage_pmax','genus', 'sample', 'gc_gc')]
# Fit the model using nls
fit_pmax_trunc <- nls(pydamage_pmax  ~ a+b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_pmax[plottable_pmax$median_age_yrs_bf_1950<32000,],
start = list(r = 1e-4, b=0.6,a=0.2))
summary(fit_pmax_trunc)

```

### 3.2.2 Confidence and Prediction intervals
```{r,message=FALSE}
# Prediction intervals
#full data
CI_pmax_full <- predictNLS(fit_pmax_full,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_pmax$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
#truncated data
CI_pmax_trunc <- predictNLS(fit_pmax_trunc,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_pmax$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary

```

### 3.2.3 Plot
```{r}
plottable_pmax$CI_lower_full <- CI_pmax_full[, "Sim.2.5%"]
plottable_pmax$CI_upper_full <- CI_pmax_full[, "Sim.97.5%"]
plottable_pmax$fitted_full <- predict(fit_pmax_full)
plottable_pmax$residual_full=residuals(fit_pmax_full)

plottable_pmax$CI_lower_trunc <- CI_pmax_trunc[, "Sim.2.5%"]
plottable_pmax$CI_upper_trunc <- CI_pmax_trunc[, "Sim.97.5%"]
plottable_pmax$fitted_trunc = predict(fit_pmax_trunc, newdata = plottable_pmax)
plottable_pmax$residual_trunc= plottable_pmax$pydamage_pmax - predict(fit_pmax_trunc, newdata = plottable_pmax)

```


```{r, fig.width=8, fig.height=5 }
p_pmax=plot_saturation_curves(plottable_pmax, pydamage_pmax)
p_pmax
```


### 3.2.5 Residuals
```{r, fig.width=9}
pmax_res=plot_residuals(plottable_pmax)
pmax_res
```

## 3.3 p_geom over time
### 3.3.1 Fit 
```{r}
plottable_pgeom <- data_w[,c('median_age_yrs_bf_1950', 'compute_length_geom_p', 'genus', 'sample','gc_gc')]
```

```{r}
fit_pgeom_full<- nls(compute_length_geom_p ~ a+ b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_pgeom,
start = list(b = 0.12, r = 0.0001, a=0.02),
control = nls.control(maxiter = 100, warnOnly = FALSE))
# Summary of fit
summary(fit_pgeom_full)
```
```{r}
fit_pgeom_trunc<- nls(compute_length_geom_p ~ a+ b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_pgeom[plottable_pgeom$median_age_yrs_bf_1950 <32000,],
start = list(b = 0.12, r = 0.0001, a=0.02),
control = nls.control(maxiter = 100, warnOnly = FALSE))
summary(fit_pgeom_trunc)
```

### 3.3.2 Confidence and Predcition intervals
```{r,message=FALSE}
# Prediction intervals
#full data
CI_pgeom_full <- predictNLS(fit_pgeom_full,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_pgeom$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
#truncated data
CI_pgeom_trunc <- predictNLS(fit_pgeom_trunc,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_pgeom$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
```

### 3.3.3 Plot
```{r}
plottable_pgeom$CI_lower_full <- CI_pgeom_full[, "Sim.2.5%"]
plottable_pgeom$CI_upper_full <- CI_pgeom_full[, "Sim.97.5%"]
plottable_pgeom$fitted_full <- predict(fit_pgeom_full)
plottable_pgeom$residual_full=residuals(fit_pgeom_full)

plottable_pgeom$CI_lower_trunc <- CI_pgeom_trunc[, "Sim.2.5%"]
plottable_pgeom$CI_upper_trunc <- CI_pgeom_trunc[, "Sim.97.5%"]
plottable_pgeom$fitted_trunc = predict(fit_pgeom_trunc, newdata = plottable_pgeom)
plottable_pgeom$residual_trunc= plottable_pgeom$compute_length_geom_p - predict(fit_pgeom_trunc, newdata = plottable_pgeom)
```
```{r}
p_pgeom=plot_saturation_curves(plottable_pgeom, compute_length_geom_p)
p_pgeom
```

### 3.3.5 Residuals

```{r, fig.width=8}
p_geom_res=plot_residuals(plottable_pgeom)
p_geom_res
```

## 3.4 length mode over time
### 3.4.1 Fit
```{r}
plottable_mode <- data_w[,c('median_age_yrs_bf_1950', 'compute_length_mode', 'genus', 'sample', 'gc_gc')]
```
```{r}
fit_mode_full <- nls(
compute_length_mode ~ Lmin + (L0 - Lmin) * exp(-lambda * median_age_yrs_bf_1950),
data = plottable_mode,
start = list(L0 = 70, Lmin = 30, lambda = 0.0001))

summary(fit_mode_full)

```
```{r}
fit_mode_trunc <- nls(
compute_length_mode ~ Lmin + (L0 - Lmin) * exp(-lambda * median_age_yrs_bf_1950),
data = plottable_mode[plottable_mode$median_age_yrs_bf_1950<32000,],
start = list(L0 = 70, Lmin = 30, lambda = 0.0001))

summary(fit_mode_trunc)
```

### 3.4.2 Confidence and Prediction Intervals
```{r,message=FALSE}
# Prediction intervals
#full data
CI_mode_full <- predictNLS(fit_mode_full,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_mode$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
#truncated data
CI_mode_trunc <- predictNLS(fit_mode_trunc,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_mode$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
```

### 3.4.3 Plot
```{r}
plottable_mode$CI_lower_full <- CI_mode_full[, "Sim.2.5%"]
plottable_mode$CI_upper_full <- CI_mode_full[, "Sim.97.5%"]
plottable_mode$fitted_full <- predict(fit_mode_full)
plottable_mode$residual_full=residuals(fit_mode_full)

plottable_mode$CI_lower_trunc <- CI_mode_trunc[, "Sim.2.5%"]
plottable_mode$CI_upper_trunc <- CI_mode_trunc[, "Sim.97.5%"]
plottable_mode$fitted_trunc = predict(fit_mode_trunc, newdata = plottable_mode)
plottable_mode$residual_trunc= plottable_mode$compute_length_mode - predict(fit_mode_trunc, newdata = plottable_mode)

```
```{r}
p_mode = plot_saturation_curves(plottable_mode, compute_length_mode)
  
p_mode
```

### 3.4.5 Residuals
```{r}
mode_res=plot_residuals(plottable_mode)
mode_res
```

## 3.5 1/lambda over age - mean overhang length
### 3.5.1 Fit
```{r}
plottable_overhangs <- data_w[,c('median_age_yrs_bf_1950', 'mapdamage_lambda', 'genus', 'sample', 'gc_gc')]
```
```{r}
fit_overhangs_full <- nls(1/mapdamage_lambda ~ a+ b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_overhangs,
start = list(b = 2, r = 0.0001, a=2),
control = nls.control(maxiter = 100, warnOnly = TRUE))

summary(fit_overhangs_full)
```
```{r}
fit_overhangs_trunc <- nls(1/mapdamage_lambda ~ a+ b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_overhangs[plottable_overhangs$median_age_yrs_bf_1950<32000,],
start = list(b = 2, r = 0.0001, a=2),
control = nls.control(maxiter = 100, warnOnly = TRUE))

summary(fit_overhangs_trunc)
```

### 3.5.2 Confidence and prediction intervals
```{r,message=FALSE}
# Prediction intervals
#full data
CI_overhangs_full <- predictNLS(fit_overhangs_full,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_overhangs$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
#truncated data
CI_overhangs_trunc <- predictNLS(fit_overhangs_trunc,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_overhangs$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
```

### 3.5.3 Plot
```{r}
plottable_overhangs$CI_lower_full <- CI_overhangs_full[, "Sim.2.5%"]
plottable_overhangs$CI_upper_full <- CI_overhangs_full[, "Sim.97.5%"]
plottable_overhangs$fitted_full <- predict(fit_overhangs_full)
plottable_overhangs$residual_full=residuals(fit_overhangs_full)

plottable_overhangs$CI_lower_trunc <- CI_overhangs_trunc[, "Sim.2.5%"]
plottable_overhangs$CI_upper_trunc <- CI_overhangs_trunc[, "Sim.97.5%"]
plottable_overhangs$fitted_trunc = predict(fit_overhangs_trunc, newdata = plottable_overhangs)
plottable_overhangs$residual_trunc= 1/plottable_overhangs$mapdamage_lambda - predict(fit_overhangs_trunc, newdata = plottable_overhangs)
```
```{r}
p_lambda=plot_saturation_curves(plottable_overhangs, 1/mapdamage_lambda)
  
 
p_lambda
```


### 3.5.5 Residuals
```{r}
lambda_res=plot_residuals(plottable_overhangs)
lambda_res
```

# 4. Fit logit model for probabilisitic age ranking based on p_max_differences
## 4.1 constructing paired datasets
#### Training data
```{r}
pairs_train = data_w[data_w$median_age_yrs_bf_1950<32000,] %>%
  dplyr::select(sample,median_age_yrs_bf_1950, pydamage_pmax) %>%
  rename_with(~ paste0(.x, "_i")) %>%
  crossing(
    data_w[data_w$median_age_yrs_bf_1950<32000,] %>%
      dplyr::select(sample, median_age_yrs_bf_1950, pydamage_pmax) %>%
      rename_with(~ paste0(.x, "_j"))) %>%
  filter(sample_i != sample_j) %>%
  mutate(
    delta_pmax = pydamage_pmax_j - pydamage_pmax_i,
    y_ij = as.integer(median_age_yrs_bf_1950_j > median_age_yrs_bf_1950_i))
```
```{r, fig.width=3.5}
hist=ggplot(pairs_train, aes(x=delta_pmax, fill=as.factor(y_ij)))+
  geom_histogram(alpha=0.2,position = 'identity', binwidth=0.01)+
  theme_bw(base_size = 9)+
  scale_fill_manual(values=c('gray40','darkred'))+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(expand=c(0,0))+
  labs(x = expression(p[max*','*j]*' - '*p[max*','*i] ), y='Count')+
  theme(legend.position='none')+
  annotate('text', x = 0.3, y = 1500, label = expression(Age[j]*'>'*Age[i]), size=3, color='darkred')+
  annotate('text', x = -0.3, y = 1500, label = expression(Age[j]*'<'*Age[i]), size=3, color='gray40')
hist
```

#### Test data
```{r}
pairs_test <- test_genera_w %>%
  dplyr::select(sample, median_age_yrs_bf_1950, pydamage_pmax) %>%
  rename_with(~ paste0(.x, "_i")) %>%
  tidyr::crossing(
    test_genera_w %>%
      dplyr::select(sample, median_age_yrs_bf_1950, pydamage_pmax) %>%
      rename_with(~ paste0(.x, "_j"))) %>%
  filter(sample_i != sample_j) %>%
  mutate(
    y_ij = as.integer(median_age_yrs_bf_1950_j > median_age_yrs_bf_1950_i),
    delta_pmax = pydamage_pmax_j - pydamage_pmax_i)
```

## 4.2 fitting logit model 
```{r}
pmax_logit = glm(
  y_ij ~ delta_pmax,
  family = binomial(),
  data = pairs_train
)
summary(pmax_logit)
```

### 4.2.1 bootstrapping confidence band
```{r}
bootstrap_pairwise_by_genus_sample <- function(
  data,
  time_col = "median_age_yrs_bf_1950",
  damage_col = "pydamage_pmax",
  sample_col = "sample",
  genus_col = "genus",
  R = 100,
  seed = 1963) {
  set.seed(seed)
 #construct pairs from data
  all_pairs <- data %>%
    rename_with(~ paste0(.x, "_i")) %>%
    crossing(
      data %>% rename_with(~ paste0(.x, "_j"))
    ) %>%
    filter(!!sym(paste0(sample_col, "_i")) != !!sym(paste0(sample_col, "_j"))) %>%
    mutate(delta_pmax = !!sym(paste0(damage_col, "_j")) - !!sym(paste0(damage_col, "_i")))
#define range of delta_pmax  
  delta_seq <- seq(
    min(all_pairs$delta_pmax),
    max(all_pairs$delta_pmax),
    length.out = 100)
  # Precompute unique levels and their row indices
  genus_levels = unique(data[[genus_col]])
  sample_levels = unique(data[[sample_col]])
  genus_index_list = lapply(genus_levels, function(g) which(data[[genus_col]]  == g))
  sample_index_list = lapply(sample_levels, function(s) which(data[[sample_col]] == s))
  # Container for predictions
  pred_matrix <- matrix(NA, nrow = R, ncol = length(delta_seq))
  for (r in seq_len(R)) {
    ## 1) Resample genus *clusters* by index, with replacement
    g_ids = sample(seq_along(genus_levels),
                    size = length(genus_levels),
                    replace = TRUE)
    ## 2) Resample sample *clusters* by index, with replacement
    s_ids = sample(seq_along(sample_levels),
                    size = length(sample_levels),
                    replace = TRUE)
    ## 3) Get the row indices for the resampled clusters
    rows_g = unlist(genus_index_list[g_ids])
    rows_s = unlist(sample_index_list[s_ids])
    ## 4) Intersection: rows that belong to both a resampled genus and sample
    rows_keep = intersect(rows_g, rows_s)
    d_resample = data[rows_keep, , drop = FALSE]
    if (nrow(d_resample) < 2) {
      next  # skip iteration if not enough data
    }
    #build pairs and fit glm based on resampled data
    pairs = d_resample %>%
      rename_with(~ paste0(.x, "_i")) %>%
      tidyr::crossing(
        d_resample %>% rename_with(~ paste0(.x, "_j"))
      ) %>%
      dplyr::filter(.data[[paste0(sample_col, "_i")]] != .data[[paste0(sample_col, "_j")]]) %>%
      dplyr::mutate(
        delta_pmax = .data[[paste0(damage_col, "_j")]] -.data[[paste0(damage_col, "_i")]],
        y_ij = as.integer(.data[[paste0(time_col, "_j")]] >.data[[paste0(time_col, "_i")]]))
    if (nrow(pairs) == 0) {
      next}
    mod = glm(y_ij ~ delta_pmax, family = binomial(), data = pairs)
    pred_matrix[r, ] = predict(mod,newdata = data.frame(delta_pmax = delta_seq),type = "response") }
# Compute 95% CI from pred_matrix
ci = apply(pred_matrix, 2, function(x) quantile(x, c(0.025, 0.975), na.rm = TRUE))
ci_df = data.frame(
  delta_pmax = delta_seq,
  lwr = ci[1, ],
  upr = ci[2, ],
  fit = apply(pred_matrix, 2, mean, na.rm = TRUE))
# Return complete results
list(bootstrap_predictions = pred_matrix,ci_df = ci_df)}
```
```{r}
result_two_way = bootstrap_pairwise_by_genus_sample(data = data_w[data_w$median_age_yrs_bf_1950<32000,] ,R = 500)
```

### 4.2.2 Plot
```{r}
plottable_logit=result_two_way$ci_df %>% mutate(orig_pred = predict(pmax_logit, 
                    newdata = data.frame(delta_pmax = .$delta_pmax), 
                    type = "response"))
```
```{r}
find_cutoff_crossing <- function(df, p) {
  df %>%
    filter(orig_pred >= p | lwr >= p | upr >= p) %>%
    summarise(
      prob = p,
      delta_pmax_fit = min(delta_pmax[orig_pred >= p], na.rm = TRUE),
      delta_pmax_lwr = min(delta_pmax[lwr >= p], na.rm = TRUE),
      delta_pmax_upr = min(delta_pmax[upr >= p], na.rm = TRUE)
    )
}

cut_095 <- find_cutoff_crossing(plottable_logit, 0.95)
cut_075 <- find_cutoff_crossing(plottable_logit, 0.75)

bind_rows(cut_095, cut_075)
```

```{r}
Scurve=ggplot(plottable_logit, aes(x = delta_pmax)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = "lightblue3", alpha = 0.5) +
  geom_line(aes(x = delta_pmax, y = orig_pred), linewidth=0.5)+
  labs(x = expression(p[max*','*j]*' - '*p[max*','*i] ), y = expression('P('*Age[j]*'>'*Age[i]*')')) +
  scale_x_continuous(expand=c(0,0))+
  theme_bw(base_size = 9)
Scurve
```

## 4.3 Application of logit-model
```{r}
pairs_test$pred_prob <- predict(
  pmax_logit ,
  newdata = pairs_test,
  type = "response"
)
```

### 4.3.1 ROC
```{r}
roc_obj <- roc(
  response = pairs_test$y_ij,
  predictor = pairs_test$pred_prob
)
```
```{r}
auc_value <- auc(roc_obj)
auc_value
```
```{r}
roc_df <- data.frame(
  specificity = roc_obj$specificities,
  sensitivity = roc_obj$sensitivities
) %>%
  mutate(
    fpr = 1 - specificity
  )

roc=
  ggplot(roc_df, aes(x = fpr, y = sensitivity)) +
  geom_line(color = "black", size = 1) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50") +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate") +
  theme_bw(base_size = 9)+
  annotate( "text",x = 0.6, y = 0.2,label = paste0("AUC = ", round(auc_value, 3)),size = 3)+
  coord_fixed()
roc
```

### 4.3.2 Classification threshold
```{r}
coords(roc_obj, "best", ret = c("threshold","sensitivity", "specificity"), best.method="youden")
```

```{r}
qlogis(c(0.5)) / coef(pmax_logit)['delta_pmax']
```
0.5 is the most suited threshold for class separation, this corresponds to a delta_pmax of 0 (as expected for logit model)

```{r}
pairs_test$pred_class <- pairs_test$pred_prob > 0.5
```

```{r}
table(
  truth = pairs_test$y_ij,
  pred  = pairs_test$pred_class
)

mean(pairs_test$y_ij == pairs_test$pred_class)
```

```{r}
mean((pairs_test$delta_pmax > 0) == pairs_test$y_ij)
```

### 4.3.3 Binned data approach
```{r}
pairs_test_plot <- pairs_test %>%
  mutate( correct = as.integer((delta_pmax > 0) == (median_age_yrs_bf_1950_j>median_age_yrs_bf_1950_i)), abs_dp = abs(delta_pmax),
      dp_bin = cut_width(abs_dp, width = 0.025, boundary = 0),include.lowest = TRUE)%>% 
 mutate(correct = factor(correct, levels = c(0, 1),
                     labels = c("Incorrect", "Correct"))) %>% 
  group_by(dp_bin, correct) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(dp_bin) %>%
  mutate(prop = n / sum(n))
```
```{r}
bars=ggplot(pairs_test_plot, aes(x = dp_bin, y = prop, fill = correct)) +
  geom_col(width = 0.9, alpha=0.5) +
  labs(
    x = expression("|"*Delta*"("*p[max]*")|"),
    y = "Probability of correct ranking",
    fill = NULL
  ) +
  scale_fill_manual(
    values = c("Incorrect" = "darkred", "Correct" = "grey40")
  ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),legend.position='none'
  )
bars
```


# 5. combined plots
### 5.1 Figure S8

```{r, fig.height=8, fig.width=7}
wrap_plots(p_deltas+ylab(expression(delta[s])), p_pmax+ylab(expression(p[max])), p_mode+ylab('Mode Fragment Length (bp)'), p_pgeom+ylab(expression(p[geom])), p_lambda+ ylab(expression('Mean Overhang Lenght '* 1/lambda *' (bp)')), ncol = 2)+  plot_annotation(tag_levels = "A")
```

## 5.2 Figure S9
```{r, fig.width= 5, fig.height= 8}
(delta_s_res+
  theme(axis.title.x = element_blank())+
  labs(title= "Residuals", subtitle = expression(delta[s]*' vs. Age Model')))/ 
(pmax_res+
   theme(axis.title.x = element_blank())+
  labs(subtitle = expression(p[max]*' vs. Age Model')))/
(mode_res+
   theme(axis.title.x = element_blank())+
  labs(subtitle = 'Mode Read Length vs. Age Model'))/ 
(p_geom_res+
   theme(axis.title.x = element_blank())+
  labs(subtitle = expression(p[geom]*' vs. Age Model')))/ 
(lambda_res+
  labs(subtitle = expression(1/lambda*' vs. Age Model')) )+ 
  plot_annotation(tag_levels = "A")+ plot_layout(guides = "collect")&theme(legend.position = "bottom",  plot.margin = margin(0, 0, 0,0)) 


 
```











