---
title: "aDNA damage analysis"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
# 1. Preparations
# 1.1 Packages
```{r}
library(scales)
packageVersion('scales')
library(tidyverse)
packageVersion('tidyverse')
library(patchwork)
packageVersion('patchwork')
library(ggrepel)
packageVersion('ggrepel')
library(propagate)
packageVersion('propagate')
library(pROC)
packageVersion('pROC')
library(ppcor)
packageVersion('ppcor')
```
# 1.2 Read the data

I worked with a training data set (data) to explore trends and fit models and a test dataset (test_genera) to test the models' performance

#### Training data
```{r}
data=read.csv('../00_input/damage_data_core_genera.csv')
```
```{r}
data_w = pivot_wider(data[,-c(3:4)],
           names_from = tool_param,
           values_from = value )
data_w$mapdamage_1_over_lambda=1/data_w$mapdamage_lambda
data_w$pydamage_1over_p=1/data_w$pydamage_p
data_w$qPCRcopies_per_uL_and_mg_rock=data_w$qPCRcopies_per_uL/data_w$mass_mg
```
#### Test data
```{r}
test_genera = read_csv("../00_input/damage_data_test_genera.csv")
```
```{r}
test_genera_w = pivot_wider(test_genera[,-c(3:4)],
           names_from = tool_param,
           values_from = value )
test_genera_w$qPCRcopies_per_uL_and_mg_rock=test_genera_w$qPCRcopies_per_uL/test_genera_w$mass_mg
```


# 2.Characterization of Pseudobins 
## 2.1 Assembly length in pseudo-bins
```{r, fig.width=6.5, fig.height=6}

# Calculate sample size per genus
n_data <- data_w %>%
  group_by(genus) %>%
  summarise(N = n())

# Add the sample size as text labels
assembly <- ggplot(data_w,
                   aes(x = factor(genus, levels = c('nitrospira',
                                                    'candidatus_nitrosotenuis',
                                                    'conexibacter',
                                                    'candidatus_methylomirabilis',
                                                    'piscinibacter',
                                                    'nitrosotalea')),
                       y = pydamage_len)) +
  geom_boxplot() +
  geom_jitter(size = 0.5, color = 'gray40', alpha=0.5) +
  scale_y_log10(limits = c(1, 10000000)) +
  theme_bw(base_size = 9) +
  coord_flip(clip = "off") +
  ylab('Assembly Length (bp)') +
  xlab('') +
  theme(axis.text.x = element_text(angle=45, hjust=1, vjust=1),
        axis.text.y = element_text(face = "italic"),plot.margin = margin(1, 1, 1, 1, "cm")) +
  scale_x_discrete(labels = c(
    "candidatus_methylomirabilis" = "Cand. Methylomirabilis\n(N=55)" , 
    "candidatus_nitrosotenuis"  = "Cand. Nitrosotenuis\n(N=57)",
    "conexibacter"  =  "Conexibacter\n(N=56)",
    "nitrosotalea"  ="Nitrosotalea\n(N=57)",
    "nitrospira"="Nitrospira\n(N=60)",
    "piscinibacter" = "Piscinibacter\n(N=56)"
  )) +
   geom_text(
    data = n_data,
    aes(
      x = factor(genus, levels = c(
        "nitrospira",
        "candidatus_nitrosotenuis",
        "conexibacter",
        "candidatus_methylomirabilis",
        "piscinibacter",
        "nitrosotalea"
      )),
      y = 1,                    # position on log10 y-axis
      label = paste0("n = ", N)
    ),
    inherit.aes = FALSE,
    hjust = 0,
    size = 2.5
  ) 
  
assembly
```
## 2.2 GC content
```{r}

gc_over_time <- data_w %>%
group_by(genus) %>%
  mutate(weight_len = log1p(pydamage_len),  # log(1 + len) to handle zeros safely
         weight_len = weight_len / max(weight_len, na.rm = TRUE)) %>%
do({
fit <- lm(gc_gc ~ median_age_yrs_bf_1950, data = .,weights = weight_len)
s <- summary(fit)
# slope = coefficient for age
slope <- coef(s)["median_age_yrs_bf_1950", "Estimate"]
se    <- coef(s)["median_age_yrs_bf_1950", "Std. Error"]
tval  <- coef(s)["median_age_yrs_bf_1950", "t value"]
pval  <- coef(s)["median_age_yrs_bf_1950", "Pr(>|t|)"]
r2    <- s$r.squared
adjr2 <- s$adj.r.squared
data.frame(
slope = slope,
se = se,
tval = tval,
pval = pval,
r2 = r2,
adjr2 = adjr2
)
}) %>%
ungroup() %>%
mutate(p_adj = p.adjust(pval, method = "BH"))
print(gc_over_time)
```
```{r}
#gc_values from literature (min/max)
gc_lit=data.frame(
  piscinibacter=c(0.662, 0.699),
  conexibacter=c(0.7, 0.74),
  nitrosotalea=c(0.371, 0.374),
  nitrospira=c(0.483,0.637),
  candidatus_nitrosotenuis=c(0.41,0.423),
  candidatus_methylomirabilis=c(0.5858, 0.5987)) %>% t() %>% data.frame() %>% rownames_to_column("genus") %>%  mutate(genus_f = factor(genus, levels=c("candidatus_nitrosotenuis","nitrosotalea" , "candidatus_methylomirabilis", "nitrospira" , "piscinibacter", "conexibacter")))
gc_lit
```

```{r, fig.width=7, fig.height=3}
genus_labels <- c(
  "candidatus_methylomirabilis" = "italic('Cand. Methylomirabilis')",
  "candidatus_nitrosotenuis"    = "italic('Cand. Nitrosotenuis')",
  "conexibacter"                = "italic('Conexibacter')",
  "nitrosotalea"                = "italic('Nitrosotalea')",
  "nitrospira"                  = "italic('Nitrospira')",
  "piscinibacter"               = "italic('Piscinibacter')"
)

gc_n <- data_w %>%mutate(genus_f = factor( genus,levels = c("candidatus_nitrosotenuis","nitrosotalea","candidatus_methylomirabilis","nitrospira","piscinibacter","conexibacter"))) %>%
count(genus_f)

gc_plot=data_w %>%
  mutate(genus_f = factor(genus, levels=c("candidatus_nitrosotenuis","nitrosotalea" , "candidatus_methylomirabilis", "nitrospira" , "piscinibacter", "conexibacter"))) %>% 
 ggplot(aes(x = '', y = gc_gc)) +
geom_hline(data = gc_lit, aes(yintercept=X1),color='lightblue3') +
geom_hline(data = gc_lit, aes(yintercept=X2),color='lightblue3') +  
   geom_rect(data = gc_lit, aes(xmin = -Inf, xmax = Inf, ymin = X1, ymax = X2),
            inherit.aes = FALSE, fill = "lightblue3", alpha = 0.3)+
geom_violin(fill='gray80', linewidth=0.2)+
geom_boxplot(width=0.2)+
facet_wrap(~ genus_f, labeller = labeller(
               genus_f = as_labeller(genus_labels, label_parsed)), nrow=1, ncol=6,drop = FALSE) +  
labs(x = "",y = "GC content (mol%)") +
theme_bw(base_size = 9)+
  scale_y_continuous(labels = percent_format(accuracy = 1))+
  geom_text(data = gc_n,aes( x = 1, y = Inf,label = paste0("italic(n) == ", n)),vjust = 1.5,size = 2.5, parse = TRUE) 
gc_plot


```

There is no significant temporal trend in GC content, except for Conexibacter, where the GC content is weakly falling with increasing age.

For genera with low to mid GC-content according to literature, the observed GC content in the aDNA pseudo-bins was higher than the literature value  

# 3. Correlation analysis
## 3.1 Correlation of parameters with age
I correlated all damage parameters with age but separated by genus to see if the trends are consistent.
I chose Spearman correlation because the expected relationship is non-linear.

```{r}
cor_data = data[data$tool_param %in% c("pydamage_pmin","pydamage_pmax","pydamage_p" ,"mapdamage_delta_d",     "mapdamage_delta_s","mapdamage_lambda","compute_length_mode", "compute_length_geom_p"),] %>%
  group_by(genus, tool_param) %>%
  summarise(
    cor_test = list(cor.test(median_age_yrs_bf_1950,  if(tool_param[1] %in% c("")) 1 / value else value, method = "spearman")),
    .groups = 'drop'
  ) %>%
  mutate(
    estimate = map_dbl(cor_test, ~ .x$estimate),
    p_value = map_dbl(cor_test, ~ .x$p.value)
  )
```

I'm adjusting p-values for multiple testing with the Banjamini-Hochberg method and create labels for plotting
```{r}
cor_data = cor_data %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  mutate(
    label = sprintf("rho = %.2f\nadj.p = %.3g", estimate, p_adj),
    stars = case_when(
    p_adj <= 0.001 ~ "***",
    p_adj <= 0.01  ~ "**",
    p_adj <= 0.05  ~ "*",
    TRUE          ~ ""),
    label_short = paste0(round(estimate, 2), stars))
```
####  Heatmap
```{r, fig.width=6, fig.height=3}
cor= cor_data %>% mutate(class=sub("_.*", "", tool_param)) %>%  ggplot(aes(y=factor(genus, levels=c('nitrospira','candidatus_nitrosotenuis','conexibacter','candidatus_methylomirabilis', 'piscinibacter','nitrosotalea')), x=factor(tool_param,levels=c('compute_length_geom_p','compute_length_mode','mapdamage_delta_s','mapdamage_delta_d','mapdamage_lambda','pydamage_pmax','pydamage_pmin','pydamage_p')), fill=estimate))+
  geom_tile()+
  scale_fill_gradient2(low = "lightblue3", mid = "white", high = "darkred", midpoint = 0, limits = c(-1, 1))+
  geom_text(aes(label = label_short), color = "black", size = 2.5)+
  scale_x_discrete(labels = c("pydamage_p" = expression(italic(p)) , 
                               "pydamage_pmin" = expression(italic(p)[min]),
                                "mapdamage_delta_s" = expression(italic(delta)[s]),
                               "compute_length_mode" ="Mode",
                               "mapdamage_lambda"=expression(italic(lambda)),
                               "mapdamage_delta_d" = expression(italic(delta)[d]),
                               "compute_length_geom_p" =expression(italic(p)[geom]),
                               "pydamage_pmax"=expression(italic(p)[max])))+
  scale_y_discrete(labels = c("candidatus_methylomirabilis" = "Cand. Methylomirabilis" , 
                              "candidatus_nitrosotenuis"  = "Cand. Nitrosotenuis",
                               "conexibacter"  =  "Conexibacter",
                               "nitrosotalea"  ="Nitrosotalea",
                               "nitrospira"="Nitrospira",
                               "piscinibacter" = "Piscinibacter"))+
   facet_grid(. ~ class, scales="free_x", space="free", labeller=labeller(class=c('compute'='Fragmentation', 'mapdamage'='mapDamage', 'pydamage'='PyDamage')))+
  theme_minimal(base_size = 9)+
  theme( legend.position = 'bottom', axis.text.y=element_text(face = "italic"),  legend.key.width = unit(1, "cm"))+
   labs(
    y = 'Genus',
    x = 'Damage Parameter',
    fill = expression(paste("Correlation with age (Spearman's ", italic(rho), ")")))
cor
```

#### Scatterplot
```{r, fig.height=8, fig.width=7}
#Helper function for the labels
tool_param_labels <- c(
  "pydamage_p" = "'PyDamage'~italic(p)",
  "pydamage_pmin" = "'PyDamage'~italic(p)[min]",
  "mapdamage_delta_s" = "'mapDamage'~italic(delta)[s]",
  "compute_length_mode" = "'Read length'~Mode",
  "mapdamage_lambda" = "'mapDamage'~italic(lambda)",
  "mapdamage_delta_d" = "'mapDamage'~italic(delta)[d]",
  "compute_length_geom_p" = "'Read length'~italic(p)[geom]",
  "pydamage_pmax" = "'PyDamage'~italic(p)[max]"
)

# Named vector of labels for genus (plain text)
genus_labels <- c(
  "candidatus_methylomirabilis" = "italic('Cand. Methylomirabilis')",
  "candidatus_nitrosotenuis"    = "italic('Cand. Nitrosotenuis')",
  "conexibacter"                = "italic('Conexibacter')",
  "nitrosotalea"                = "italic('Nitrosotalea')",
  "nitrospira"                  = "italic('Nitrospira')",
  "piscinibacter"               = "italic('Piscinibacter')"
)

scatter_cor=
data[data$tool_param %in% c("pydamage_pmin","pydamage_pmax","pydamage_p" ,"mapdamage_delta_d",     "mapdamage_delta_s","mapdamage_lambda","compute_length_mode", "compute_length_geom_p"),] %>% mutate(
    value_plot = ifelse(tool_param %in% c(""), 1 / value, value)
  ) %>% 
ggplot(aes(x = median_age_yrs_bf_1950, y = value_plot, color = genus)) +
  geom_vline(xintercept=32000, color="gray75")+
  geom_point(shape=4) +
 facet_grid(rows = vars(tool_param), cols = vars(genus),
             scales = "free_y",
             labeller = labeller(
               tool_param = as_labeller(tool_param_labels, label_parsed),
               genus = as_labeller(genus_labels, label_parsed)
             )) +
  theme_bw(base_size = 8) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),legend.position = 'bottom') +
  geom_text(size=3,
    data = cor_data,
    aes(label = label),
    x = Inf, y = Inf,
    hjust = 1.1, vjust = 1.1,
    inherit.aes = FALSE
  )+
  xlab('Nominal Age (Years BP)')+
  ylab('')+
  scale_color_manual("", values =c( "darkred","lightblue3",'#009E73',"darkorchid", "blue4", '#E66101'),guide = guide_legend(ncol = 6),   labels = parse(text = genus_labels)
) 
scatter_cor
```

Beyond 32000 years temporal trends reversed or became unstable in multiple parameters and genera. This conincides with a change in the microfabric of TM-17. Therefore we restrict the damage modelling to samples younger than 32000y.
### Influence of Reference length
```{r}
damage_params <- c(
  "pydamage_pmin",
  "pydamage_pmax",
  "pydamage_p",
  "mapdamage_delta_d",
  "mapdamage_delta_s",
  "mapdamage_lambda",
  "compute_length_mode",
  "compute_length_geom_p"
)

wide_data <- data %>%
  filter(tool_param %in% c(damage_params, "pydamage_len")) %>%
  dplyr::select(
    sample_name,
    median_age_yrs_bf_1950,
    tool_param,
    value,
    genus
  ) %>%
  pivot_wider(
    names_from = tool_param,
    values_from = value
  ) %>%
  mutate(
    across(
      all_of(c(damage_params, "pydamage_len")),
      as.numeric
    )
  )

global_cor_data <- map_dfr(
  damage_params,
  function(param) {

    spearman_test <- cor.test(
      x = wide_data$median_age_yrs_bf_1950,
      y = wide_data[[param]],
      method = "spearman"
    )

    partial_test <- pcor.test(
      x = wide_data$median_age_yrs_bf_1950,
      y = wide_data[[param]],
      z = wide_data$pydamage_len,
      method = "spearman"
    )

    tibble(
      tool_param = param,

      spearman_rho = unname(spearman_test$estimate),
      spearman_p = spearman_test$p.value,

      partial_rho = partial_test$estimate,
      partial_p = partial_test$p.value,

      delta_rho = partial_test$estimate - spearman_test$estimate
    )

  }
)
```

```{r, fig.width=7, fig.height=3}
plot_data <- global_cor_data %>%
  tidyr::pivot_longer(cols = c(spearman_rho, partial_rho),  names_to = "correlation_type",values_to = "rho") %>%
  mutate(correlation_type = recode(correlation_type,spearman_rho = "Spearman",partial_rho = "Partial Spearman"))

comparison= ggplot(global_cor_data, aes(y = reorder(tool_param, spearman_rho))) +
 # connecting lines
  geom_segment(aes(x = spearman_rho,xend = partial_rho,yend = reorder(tool_param, spearman_rho)),linewidth = 0.7,color = "grey60") +
# standard Spearman: filled black points
  geom_point( aes(x = spearman_rho, shape = "Spearman"),size = 3,color = "black",fill = "black") +
# partial Spearman: open red circles
  geom_point(aes(x = partial_rho, shape = "Partial Spearman"),size = 3, color = "darkred", stroke = 1.2 ) +
geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
scale_shape_manual(name = NULL,values = c("Spearman" = 16,"Partial Spearman" = 21),
  labels = c("Spearman"=expression("Spearman's "~rho[age*","*damage]),
  "Partial Spearman"= expression("Spearman's "~rho[age*","*damage*"|"*reference~length]))) +
labs( x = expression(rho),y = NULL) +
theme_bw(base_size = 9) +
theme(legend.position = "bottom")+
scale_y_discrete(labels = function(x) parse(text = tool_param_labels[x]))
comparison
```

## 3.2 correlation of parameters with each other
### 3.2.1 Fragment length and 1/lambda
```{r}
cor.test(data_w$compute_length_geom_p, data_w$mapdamage_1_over_lambda, method='spearman' )
```


```{r, fig.widht=8, fig.height=5}
p_cor=ggplot(data_w, aes(y=mapdamage_1_over_lambda, x=compute_length_geom_p))+
  geom_point(shape=4, color='gray40')+
  theme_bw(base_size = 9)+
   labs( y=expression("Mean Overhang length "* 1/italic(lambda)*" (bp)"),  x = expression(italic(p)[geom]))+
   annotate('text', x = 0.075, y = 2.25, label = "atop('Spearman\\'s Rho = 0.63', italic(p)*'-value < 2.2e-16')",,
  parse = TRUE, size=3)
p_cor
```

### 3.2.3 p and lambda
```{r}
cor.test(data_w$pydamage_p, data_w$mapdamage_lambda, method='spearman' )
```
```{r}
ggplot(data_w, aes(y=mapdamage_lambda, x=pydamage_p))+
  geom_point(shape=4, color='gray40')+
  theme_bw(base_size = 9)+
   labs( y=expression("mapDamage "* lambda),  x = "PyDamage p")+
   annotate('text', x = 0.6, y = 0.25,label = "atop('Spearman\\'s Rho = 0.60', italic(p)*'-value < 2.2e-16')",,
  parse = TRUE, size=2.5)
```


# 4. Exponential saturation models: Damage over time
## 4.1 delta_s over time
### 4.1.1 Fit
```{r}
plottable_delta_s=data_w[data_w$median_age_yrs_bf_1950<32000,c('median_age_yrs_bf_1950', 'mapdamage_delta_s', 'genus','sample','gc_gc','pydamage_len','pydamage_nb_contigs', 'pydamage_nb_reads')]
fit_delta_s <- nls(mapdamage_delta_s  ~ 1 - exp(-r * median_age_yrs_bf_1950),data = plottable_delta_s,start = list(r = 1e-3))
summary(fit_delta_s)
```
### 4.1.2 Confidence and Prediction interval
```{r,message=FALSE}
# Prediction interval (new observation)
PI_delta_s <- predictNLS(fit_delta_s,
                     newdata =data.frame(median_age_yrs_bf_1950=plottable_delta_s$median_age_yrs_bf_1950),interval = "prediction",alpha = 0.05,nsim = 10000)$summary
CI_delta_s <- predictNLS(fit_delta_s,newdata =data.frame(median_age_yrs_bf_1950=plottable_delta_s$median_age_yrs_bf_1950),interval = "confidence",alpha = 0.05,nsim = 10000)$summary
```

### 4.1.3 Plot
```{r}
plottable_delta_s$PI_lower <- PI_delta_s[, "Sim.2.5%"]
plottable_delta_s$PI_upper <- PI_delta_s[, "Sim.97.5%"]
plottable_delta_s$CI_lower <- CI_delta_s[, "Sim.2.5%"]
plottable_delta_s$CI_upper <- CI_delta_s[, "Sim.97.5%"]
plottable_delta_s$fitted <- predict(fit_delta_s)
plottable_delta_s$residual=residuals(fit_delta_s)

p_deltas=ggplot(plottable_delta_s, aes(x = median_age_yrs_bf_1950)) +
  geom_line(aes(y = PI_lower),,color='gray40',linetype='twodash')+
  geom_line(aes(y= pmin(pmax(PI_upper, 0), 1)),color='gray40',linetype='twodash')+
  geom_ribbon(aes(x = median_age_yrs_bf_1950,
                  ymin = CI_lower, ymax = CI_upper), fill = "lightblue3", color='lightblue3', alpha=0.5) +
  geom_point(aes(x = median_age_yrs_bf_1950, y = mapdamage_delta_s),color = "gray40", shape = 4) +
  geom_line(aes(x = median_age_yrs_bf_1950, y = fitted), color = "black") +
  labs( x = "Nominal Age (Years BP)", y = expression(italic(delta)[s]),caption= expression("Fit: "~italic(delta)[s] == 1 - e^{-italic(r) %*% Age}) )+
  geom_hline(yintercept=1, colour='darkred')+
  theme_bw(base_size=9)
  
p_deltas
```
### 4.1.4 Characteristic values

```{r}
# time to saturation (99%)
(-1/coef(fit_delta_s)['r'])*log(0.01)
```

### 4.1.5 Residuals
```{r}
delta_s_res=ggplot(plottable_delta_s, aes(x=median_age_yrs_bf_1950, y=residual, color=genus))+
geom_point(shape=4)+
geom_hline(yintercept=0)+
theme_bw(base_size = 9)+
labs(x='Nominal Age (Years BP)', y='Residual', title='Residuals', subtitle=expression(italic(delta)[s]*' vs. Age Model'))+
scale_color_manual("Genus", values =c( "darkred",   "lightblue3",    '#009E73',    "darkorchid",   "blue4",'#E66101'),
                    labels =parse(text = genus_labels))
delta_s_res
```
  
## 4.2 pmax over time
### 4.2.1 Fit
```{r}
plottable_pmax=data_w[data_w$median_age_yrs_bf_1950<32000,c('median_age_yrs_bf_1950', 'pydamage_pmax','genus', 'sample', 'gc_gc','pydamage_len','pydamage_nb_contigs', 'pydamage_nb_reads')]
# Fit the model using nls
fit_pmax <- nls(pydamage_pmax  ~ a+b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_pmax,
start = list(r = 1e-4, b=0.6,a=0.2))
summary(fit_pmax)

```
### 4.2.2 Confidence and Prediction intervals
```{r,message=FALSE}
# Confidence interval (mean curve)
CI_pmax = predictNLS(fit_pmax, newdata = data.frame(median_age_yrs_bf_1950=plottable_pmax$median_age_yrs_bf_1950 ), interval = "confidence", alpha = 0.05, nsim = 10000)$summary

# Prediction interval (new observation)
PI_pmax= predictNLS(fit_pmax, newdata =data.frame(median_age_yrs_bf_1950=plottable_pmax$median_age_yrs_bf_1950), interval = "prediction", alpha = 0.05, nsim = 10000)$summary

```

### 4.2.3 Plot

```{r, fig.width=8, fig.height=5 }
plottable_pmax$fitted <- predict(fit_pmax)
plottable_pmax$residual=residuals(fit_pmax)
plottable_pmax$CI_lower <- CI_pmax[, "Sim.2.5%"]
plottable_pmax$CI_upper <- CI_pmax[, "Sim.97.5%"]
plottable_pmax$PI_lower <- PI_pmax[, "Sim.2.5%"]
plottable_pmax$PI_upper <- PI_pmax[, "Sim.97.5%"]

p_pmax= ggplot(plottable_pmax, aes(x = median_age_yrs_bf_1950)) +
  geom_line(aes(y = PI_lower,color='95% PI'),linetype='twodash')+
  geom_line(aes(y= PI_upper,color='95% PI'),linetype='twodash')+
  geom_ribbon(aes(x = median_age_yrs_bf_1950,
                  ymin = CI_lower, ymax = CI_upper, color='95% CI' ),fill = "lightblue3", alpha=0.5) +
geom_point(aes(x = median_age_yrs_bf_1950, y = pydamage_pmax, colour='Pseudobins'), shape=4) +
geom_line(aes(x = median_age_yrs_bf_1950, y = fitted, color='Exponential Model'))+
geom_hline(aes(yintercept= coef(fit_pmax)["a"]+coef(fit_pmax)['b'], colour='Asymptote'))+
scale_color_manual(name = "",values = c('Asymptote'='darkred','Exponential Model'='black','95% PI' = 'gray40', '95% CI'='lightblue3', 'Pseudobins'='gray40' ))+
theme_bw(base_size=9)+
scale_x_continuous(expand = c(0,0), limits=c(0,32000))+
labs(
caption = expression("Fit: "~
italic(p)[max] == italic(a) + italic(b) %*% (1 - e^{italic(-r) %*% Age})) ,y=expression(italic(p)[max]), x = "Nominal Age (Years BP)")
p_pmax
```

### 4.2.4 characteristic values
```{r}
# 99% saturation time (-1/r)*log(1-0.99)
(-1/coef(fit_pmax)['r'])*log(0.01)
```
```{r}
#saturation value
coef(fit_pmax)['b']+coef(fit_pmax)["a"]
```


### 4.2.5 Residuals
```{r, fig.width=9}
pmax_res=ggplot(plottable_pmax, aes(x=median_age_yrs_bf_1950, y=residual, color=genus))+
geom_point(shape=4)+
geom_hline(yintercept=0)+
theme_bw(base_size = 9)+
labs(x='Nominal Age (Years BP)', y='Residual',subtitle=expression(italic(p)[max]*' vs. Age Model'))+
scale_color_manual("Genus", values =c( "darkred",   "lightblue3",    '#009E73',    "darkorchid",   "blue4",'#E66101'),
                  labels = parse(text = genus_labels))
pmax_res
```

## 4.3 p_geom over time
### 4.3.1 Fit
```{r}
plottable_pgeom <- data_w[data_w$median_age_yrs_bf_1950<32000,c('median_age_yrs_bf_1950', 'compute_length_geom_p', 'genus', 'sample','gc_gc','pydamage_len','pydamage_nb_contigs', 'pydamage_nb_reads')]

fit_pgeom<- nls(compute_length_geom_p ~ a+ b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_pgeom,
start = list(b = 0.12, r = 0.0001, a=0.02),
control = nls.control(maxiter = 100, warnOnly = FALSE))
# Summary of fit
summary(fit_pgeom)

```
### 4.3.2 Confidence and Predcition intervals
```{r,message=FALSE}
# Confidence interval (mean curve)
CI_pgeom = predictNLS(fit_pgeom, newdata = data.frame(median_age_yrs_bf_1950=plottable_pgeom$median_age_yrs_bf_1950 ), interval = "confidence", alpha = 0.05, nsim = 10000)$summary

# Prediction interval (new observation)
PI_pgeom= predictNLS(fit_pgeom, newdata =data.frame(median_age_yrs_bf_1950=plottable_pgeom$median_age_yrs_bf_1950), interval = "prediction", alpha = 0.05, nsim = 10000)$summary
```

### 4.3.3 Plot
```{r}
plottable_pgeom$fitted <- predict(fit_pgeom)
plottable_pgeom$residual= residuals(fit_pgeom)
plottable_pgeom$CI_lower <- CI_pgeom[, "Sim.2.5%"]
plottable_pgeom$CI_upper <- CI_pgeom[, "Sim.97.5%"]
plottable_pgeom$PI_lower <- PI_pgeom[, "Sim.2.5%"]
plottable_pgeom$PI_upper <- PI_pgeom[, "Sim.97.5%"]

  p_pgeom=ggplot(plottable_pgeom, aes(x=median_age_yrs_bf_1950))+
  geom_line(aes(y = PI_lower,color='95% PI'),linetype='twodash')+
  geom_line(aes(y= PI_upper,color='95% PI'),linetype='twodash')+
  geom_ribbon(aes(x = median_age_yrs_bf_1950,
                  ymin = CI_lower, ymax = CI_upper, color= '95% CI'), fill = "lightblue3", alpha=0.5) +
geom_point(aes(x=median_age_yrs_bf_1950, y=compute_length_geom_p,color='Pseudobins'), shape=4)+
geom_line(aes(x=median_age_yrs_bf_1950, y=fitted, color='Exponential Model'))+
geom_hline( aes(yintercept=coef(fit_pgeom)['b']+coef(fit_pgeom)['a'],colour='Asymptote'))+
theme_bw(base_size = 9)+
scale_color_manual(name = "",values = c('Asymptote'='darkred','Exponential Model'='black','95% PI' = 'gray40', '95% CI'='lightblue3', 'Pseudobins'='gray40' ))+
scale_x_continuous(expand = c(0,0), limits=c(0,32000))+
labs(
caption = expression("Fit: "~
italic(p)[geom] == italic(a) + italic(b) %*% (1 - e^{italic(-r) %*% Age})) ,y=expression(italic(p)[geom]), x = "Nominal Age (Years BP)")
p_pgeom
```

### 4.3.4 Characteristic values

```{r}
#saturation value p_inf+a
coef(fit_pgeom)['b']+coef(fit_pgeom)['a']
```

```{r}
# 99% saturation time (-1/r)*log(1-0.99)
(-1/coef(fit_pgeom)['r'])*log(0.01)
```

### 4.3.5 Residuals

```{r, fig.width=8}
p_geom_res=ggplot(plottable_pgeom, aes(x=median_age_yrs_bf_1950, y=residual, color=genus))+
geom_point(shape=4)+
geom_hline(yintercept=0)+
theme_bw(base_size = 9)+
labs(x='Nominal Age (Years BP)', y='Residual',subtitle=expression(italic(p)[geom]*' vs. Age Model'))+
scale_color_manual("Genus", values =c( "darkred",   "lightblue3",    '#009E73',    "darkorchid",   "blue4",'#E66101'),
                  labels = parse(text = genus_labels))
p_geom_res
```

## 4.4 length mode over time
### 4.4.1 Fit
```{r}
plottable_mode <- data_w[data_w$median_age_yrs_bf_1950<32000,c('median_age_yrs_bf_1950', 'compute_length_mode', 'genus', 'sample', 'gc_gc','pydamage_len','pydamage_nb_contigs', 'pydamage_nb_reads')]

fit_mode <- nls(
compute_length_mode ~ Lmin + (L0 - Lmin) * exp(-lambda * median_age_yrs_bf_1950),
data = plottable_mode,
start = list(L0 = 70, Lmin = 30, lambda = 0.0001))

summary(fit_mode)
```
### 4.4.2 Confidence and Prediction Intervals
```{r,message=FALSE}
# Confidence interval (mean curve)
CI_mode = predictNLS(fit_mode, newdata = data.frame(median_age_yrs_bf_1950=plottable_mode$median_age_yrs_bf_1950 ), interval = "confidence", alpha = 0.05, nsim = 10000)$summary

# Prediction interval (new observation)
PI_mode= predictNLS(fit_mode, newdata =data.frame(median_age_yrs_bf_1950=plottable_mode$median_age_yrs_bf_1950), interval = "prediction", alpha = 0.05, nsim = 10000)$summary
```
 ### 4.4.3 Plot
```{r}
plottable_mode$fitted <- predict(fit_mode)
plottable_mode$residual=residuals(fit_mode)
plottable_mode$CI_lower <- CI_mode[, "Sim.2.5%"]
plottable_mode$CI_upper <- CI_mode[, "Sim.97.5%"]

plottable_mode$PI_lower <- PI_mode[, "Sim.2.5%"]
plottable_mode$PI_upper <- PI_mode[, "Sim.97.5%"]

p_mode =ggplot(plottable_mode, aes(x = median_age_yrs_bf_1950)) +
   geom_line(aes(y = PI_lower),color="gray40",linetype='twodash')+
  geom_line(aes(y= PI_upper),color="gray40",linetype='twodash')+
  geom_ribbon(aes(x = median_age_yrs_bf_1950,
                  ymin = CI_lower, ymax = CI_upper), fill = "lightblue3", color='lightblue3', alpha=0.5) +
geom_point(aes(y = compute_length_mode), color = "gray40", shape=4) +
geom_line(aes(y = fitted), color = "black") +
labs(x = "Nominal Age (Years BP)", y = "Mode Fragment Length \n(bp)", caption = expression("Fit: "~
mode == italic(L)[min]+(italic(L)[0]-italic(L)[min]) %*%  e^{italic(-r) %*% Age}) ) +
geom_hline(yintercept = coef(fit_mode)['Lmin'],  color = "darkred") +
theme_bw(base_size=9)+
scale_x_continuous(expand = c(0,0), limits=c(0,32000))
p_mode
```

### 4.4.4 Characterisitc Values
```{r}
#time to reach 99% saturation
(-1/coef(fit_mode)['lambda'])*log(0.01)
```
### 4.4.5 Residuals
```{r}
mode_res=ggplot(plottable_mode, aes(x=median_age_yrs_bf_1950, y=residual, color=genus))+
geom_point(shape=4)+
geom_hline(yintercept=0)+
theme_bw(base_size = 9)+
labs(x='Nominal Age (Years BP)', y='Residual',subtitle='Mode Read Length vs. Age Model')+
scale_color_manual("Genus", values =c( "darkred",   "lightblue3",    '#009E73',    "darkorchid",   "blue4",'#E66101'),
                  labels = parse(text = genus_labels))
mode_res
 
```

## 4.5 1/lambda over age - mean overhang length
### 4.5.1 Fit
```{r}
plottable_overhangs <- data_w[data_w$median_age_yrs_bf_1950<32000,c('median_age_yrs_bf_1950', 'mapdamage_lambda', 'genus', 'sample', 'gc_gc','pydamage_len','pydamage_nb_contigs', 'pydamage_nb_reads')]

fit_overhangs <- nls(1/mapdamage_lambda ~ a+ b * (1 - exp(-r * median_age_yrs_bf_1950)),
data = plottable_overhangs,
start = list(b = 2, r = 0.0001, a=2),
control = nls.control(maxiter = 100, warnOnly = TRUE))

summary(fit_overhangs)
```
### 4.5.2 Confidence and prediction intervals
```{r,message=FALSE}
# Confidence interval (mean curve)
CI_overhangs = predictNLS(fit_overhangs, newdata = data.frame(median_age_yrs_bf_1950=plottable_overhangs$median_age_yrs_bf_1950 ), interval = "confidence", alpha = 0.05, nsim = 10000)$summary

# Prediction interval (new observation)
PI_overhangs= predictNLS(fit_overhangs, newdata =data.frame(median_age_yrs_bf_1950=plottable_overhangs$median_age_yrs_bf_1950), interval = "prediction", alpha = 0.05, nsim = 10000)$summary
```

### 4.5.3 Plot
```{r}
plottable_overhangs$fitted = predict(fit_overhangs)
plottable_overhangs$residual = residuals(fit_overhangs)
plottable_overhangs$CI_lower <- CI_overhangs[, "Sim.2.5%"]
plottable_overhangs$CI_upper <- CI_overhangs[, "Sim.97.5%"]
plottable_overhangs$PI_lower <- PI_overhangs[, "Sim.2.5%"]
plottable_overhangs$PI_upper <- PI_overhangs[, "Sim.97.5%"]


p_lambda= ggplot(plottable_overhangs,aes(x=median_age_yrs_bf_1950))+
  geom_line(aes(y = PI_lower),color='gray40',linetype='twodash')+
  geom_line(aes(y= PI_upper),color='gray40',linetype='twodash')+
  geom_ribbon(aes(x = median_age_yrs_bf_1950,
                  ymin = CI_lower, ymax = CI_upper ),fill = 'lightblue3', color='lightblue3', alpha=0.5) +
geom_point(aes(x=median_age_yrs_bf_1950, y=1/mapdamage_lambda),shape=4, color='gray40')+
geom_line(aes(x=median_age_yrs_bf_1950, y=fitted))+
geom_hline(yintercept=(coef(fit_overhangs)['b'])+coef(fit_overhangs)['a'], colour='darkred')+
theme_bw(base_size = 9)+
scale_x_continuous(expand = c(0,0), limits=c(0,32000))+
labs(
caption = expression("Fit: " ~ frac(1, italic(lambda)) == italic(a) + italic(b) %*% (1 - e^{italic(-r) %*% Age})),
y=expression("Mean Overhang length "* 1/italic(lambda)*" (bp)"), 
x = "Nominal Age (Years BP)")
p_lambda
```
### 4.5.4 Characteristic values
```{r}
#saturation value
coef(fit_overhangs)['b']+coef(fit_overhangs)['a']
```
```{r}
#time to 99% saturation
(-1/coef(fit_overhangs)['r'])*log(0.01)
```
### 4.5.5 Residuals
```{r}
lambda_res=ggplot(plottable_overhangs, aes(x=median_age_yrs_bf_1950, y=residual, color=genus))+
geom_point(shape=4)+
geom_hline(yintercept=0)+
theme_bw(base_size = 9)+
labs(x='Nominal Age (Years BP)', y='Residual', subtitle=expression(~1/italic(lambda)*' vs. Age Model'))+
scale_color_manual("Genus", values =c( "darkred",   "lightblue3",    '#009E73',    "darkorchid",   "blue4",'#E66101'),
                  labels = parse(text = genus_labels))
lambda_res
```

 ## 4.6 Screening for influence of GC content and cell wall morphology
```{r}
residuals_overview=list(plottable_delta_s,plottable_mode,plottable_overhangs,plottable_pgeom,plottable_pmax)
```

```{r}
# Function to summarize a single model data frame by genus
summarize_by_genus <- function(df) {
  df %>%
    group_by(genus) %>%
    summarise(
      median_gc = median(gc_gc, na.rm = TRUE),
      q1_gc = quantile(gc_gc, 0.25, na.rm = TRUE),
      q3_gc = quantile(gc_gc, 0.75, na.rm = TRUE),
      median_resid = median(residual, na.rm = TRUE),
      q1_resid = quantile(residual, 0.25, na.rm = TRUE),
      q3_resid = quantile(residual, 0.75, na.rm = TRUE),
      median_len=median(pydamage_len, na.rm =TRUE),
      sd_len=sd(pydamage_len, na.rm=TRUE),
      n=sum(!is.na(gc_gc)),
      .groups = 'drop'
    ) %>% 
    mutate(
      cell_wall = case_when(
        genus == "conexibacter" ~ "gram_pos",
        genus %in% c("nitrosotalea", "candidatus_nitrosotenuis") ~ "archaeon",
        TRUE ~ "gram_neg"
      ))}
# Apply the summarization to all models
genus_summary_list <- lapply(residuals_overview, summarize_by_genus)
```
```{r}
genus_labels_plain <- c(
  "candidatus_methylomirabilis" = "Cand. Methylomirabilis",
  "candidatus_nitrosotenuis"= "Cand. Nitrosotenuis",
  "conexibacter"= "Conexibacter",
  "nitrosotalea"= "Nitrosotalea",
  "nitrospira"= "Nitrospira",
  "piscinibacter"= "Piscinibacter"
)
make_genus_labels <- function(df) {
  labels <- paste0("italic('",genus_labels_plain[df$genus],"')~'('*italic(n)*' = '*",df$n,"*')'")
  setNames(labels, df$genus)}

```

```{r}
plot_genus_model = function(genus_summary_list) {
  labels = c(expression(italic(delta)[s]), "Mode Fragment Length",expression(1 / italic(lambda)),expression(italic(p)[geom]), expression(italic(p)[max]))
  
  results = vector("list", length(genus_summary_list))
  for (i in seq_along(genus_summary_list)) {
    df = genus_summary_list[[i]]
    results[[i]] = ggplot(
      df,aes(x = median_gc, y = median_resid, shape = cell_wall, color=genus, group=1)) +
       geom_smooth( method = "lm",formula = y ~ x,se = FALSE, aes(group = 1), color = "black", linewidth = 0.4)+
      geom_point(size = 2.5) +
      geom_errorbar(aes(ymin = q1_resid,  ymax = q3_resid),width = 0) +
      geom_errorbar(aes(xmin = q1_gc,xmax = q3_gc),height = 0) +
      labs(x = "Median GC content",y = "Median residual", title = labels[i]
      ) +
      theme_bw(base_size = 9)+
      #scale_color_manual('Cell Wall Morphology', values = c('gram_pos'='lightblue3','archaeon'='darkred','gram_neg'='gray40' ),labels=c('gram_pos'='gram+','gram_neg'= 'gram-', 'archaeon'='archaeal'))+
      scale_shape_manual('Cell Wall Morphology', values = c('gram_pos'=17,'archaeon'=16,'gram_neg'=4),labels=c('gram_pos'='gram+','gram_neg'= 'gram-', 'archaeon'='archaeal'))+
      scale_color_manual("Genus", values =c( "darkred",   "lightblue3",    '#009E73',    "darkorchid",   "blue4",'#E66101'),
                  labels = parse(text = make_genus_labels(df)))
  }
  return(results)
}
# Apply to all models in your list
results_list_gc = plot_genus_model(genus_summary_list)
results_list_gc[[1]]
```

## 4.7 Screening for influence of reference length
```{r}
plot_res_vs_len = function(list) {
  labels = c(expression(italic(delta)[s]), "Mode Fragment Length",expression(1 / italic(lambda)),expression(italic(p)[geom]), expression(italic(p)[max]))
  
  results = vector("list", length(list))
  for (i in seq_along(list)) {
    df = list[[i]]
    results[[i]] = ggplot(
      df,aes(y = residual, x = pydamage_len, color = genus, group=1)) +
      geom_point(size = 2, shape=4) +
      geom_smooth( method = "lm",formula = y ~ x,se = FALSE, aes(group = 1), color='black', linewidth = 0.4)+
      labs(x = "Pseudobin lenght (bp)",y = "Residual", title = labels[i]) +
      theme_bw(base_size = 9)+
     # facet_wrap(~genus)+
      scale_x_log10()+
      scale_color_manual("Genus", values =c( "darkred",   "lightblue3",    '#009E73',    "darkorchid",   "blue4",'#E66101'), labels = parse(text = genus_labels))
  }
  return(results)
}
# Apply to all models in your list
results_list_len = plot_res_vs_len(residuals_overview)
results_list_len[[1]]
results_list_len[[2]]
results_list_len[[3]]
results_list_len[[4]]
results_list_len[[5]]
```
```{r}
lm(residual ~ log(pydamage_len), data = plottable_pmax) %>% summary()
```
```{r}
lm(residual ~ log(pydamage_len), data = plottable_delta_s) %>% summary()
```
```{r}
lm(residual ~ log(pydamage_len), data = plottable_mode) %>% summary()
```
```{r}
lm(residual ~ log(pydamage_len), data = plottable_pgeom) %>% summary()
```
```{r}
lm(residual ~ log(pydamage_len), data = plottable_overhangs) %>% summary()
```




# 5. Application of p_max-age-model for absolute age prediction
```{r}
test_genera_w=test_genera_w[test_genera_w$median_age_yrs_bf_1950 <32000,]
```

## 5.1 Overview of test data

```{r}
test_genera_w$range_status <- ifelse(
  test_genera_w$pydamage_pmax < coef(fit_pmax)['a'] |
    test_genera_w$pydamage_pmax >coef(fit_pmax)['b']+coef(fit_pmax)['a'],
  "Outside defined model range",
  "Within defined model range"
)

overview <- ggplot(test_genera_w,aes(x = median_age_yrs_bf_1950,y = pydamage_pmax)) +
  #geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=model_lower,ymax=model_upper),fill="grey90") +
  geom_line(data=plottable_pmax,aes(x=median_age_yrs_bf_1950,y=fitted,linetype="Exponential model"),inherit.aes=FALSE,linewidth=0.8) +
  geom_line(data=plottable_pmax,aes(x=median_age_yrs_bf_1950,y=PI_lower,linetype="Prediction interval limits"),inherit.aes=FALSE,linewidth=0.5) +
  geom_line(data=plottable_pmax,aes(x=median_age_yrs_bf_1950,y=PI_upper,linetype="Prediction interval limits"),inherit.aes=FALSE,linewidth=0.5) +
  geom_hline(aes(yintercept=coef(fit_pmax)['a'],linetype="Defined range limits"),color="darkred") +
  geom_hline(aes(yintercept=coef(fit_pmax)['b']+coef(fit_pmax)['a'],linetype="Defined range limits"),color="darkred") +
  geom_point(aes(color=range_status),shape=4) +
  scale_color_manual(name="Overlay of test genera",values=c("Within defined model range"="grey40","Outside defined model range"="darkred")) +
  scale_linetype_manual(name="Model fitted to core genera",values=c("Exponential model"="solid","Prediction interval limits"="twodash","Defined range limits"="solid")) +
  labs(x="Nominal Age (Years BP)",y=expression(italic(p)[max])) +
  theme_bw(base_size=9) +
  theme(panel.grid.minor=element_blank(),legend.position = c(0.98, 0.02),
  legend.justification = c(1, 0),legend.box="horizontal",legend.box.just="top",legend.direction = "vertical",legend.background = element_rect(fill = NA),legend.box.background = element_rect(
    fill = "white",
    colour = "grey50",
    linewidth = 0.4
  ))+
  ylim(c(0,0.75))
overview
```
```{r}
plottable_overview <- rbind(data.frame(
    source = "test_genera_overlay",
    age = test_genera_w$median_age_yrs_bf_1950,
    pmax = test_genera_w$pydamage_pmax,
    fitted = NA,
    PI_lower = NA,
    PI_upper = NA,
    defined_lower = NA,
    defined_upper = NA,
    range_status = test_genera_w$range_status
  ),
  data.frame(
    source = "core_genera_model",
    age = plottable_pmax$median_age_yrs_bf_1950,
    pmax = NA,
    fitted = plottable_pmax$fitted,
    PI_lower = plottable_pmax$PI_lower,
    PI_upper = plottable_pmax$PI_upper,
    defined_lower = coef(fit_pmax)['a'],
    defined_upper = coef(fit_pmax)['b']+coef(fit_pmax)['a'],
    range_status = NA
  )
)
```

Pseudobins exceeding the range of the model:
```{r}
sum(test_genera_w$pydamage_pmax >=coef(fit_pmax)['b']+coef(fit_pmax)["a"]|test_genera_w$pydamage_pmax <= coef(fit_pmax)["a"] )
```


## 5.2 Model inversion and age prediction

```{r}
invert_age = function(pmax, a, b, r) {
  # Return NA if pmax outside model range
  if(pmax <= a | pmax >= (a + b)) return(NA)
  -1 / r * log(1 - (pmax - a)/b)
}
```
```{r}
predict_age_from_pmax <- function(df, pmax_col, model, B = 1000, seed = 1963) {
  set.seed(seed)
  
  # Extract parameter estimates and covariance
  params_est = coef(model)
  cov_mat = vcov(model)
  sigma_resid = summary(model)$sigma
  
  # Prepare storage vectors
  n = nrow(df)
  age_point = numeric(n)
  age_ci_lower <- numeric(n)
  age_ci_upper <- numeric(n)
  age_pi_lower = numeric(n)
  age_pi_upper = numeric(n)
  age_pi_bootstrap = vector("list", n)
  
  for(i in 1:n) {
    pmax = df[[pmax_col]][i]
#Point estimate
    age_point[i] = invert_age(pmax, params_est["a"], params_est["b"], params_est["r"])
    
#CI: parametric bootstrap without residual noise
    param_samples <- mvrnorm(B, mu = params_est, Sigma = cov_mat)
    age_boot_ci <- sapply(1:B, function(j) {
      invert_age(pmax, param_samples[j, "a"], param_samples[j, "b"], param_samples[j, "r"])
    })
    age_boot_ci <- age_boot_ci[!is.na(age_boot_ci)]
    age_ci_lower[i] <- quantile(age_boot_ci, 0.025)
    age_ci_upper[i] <- quantile(age_boot_ci, 0.975)
#PI: parametric bootstrap with residual noise
    param_samples = mvrnorm(B, mu = params_est, Sigma = cov_mat)
    age_boot_pi = sapply(1:B, function(j) {
      a_j = param_samples[j, "a"]
      b_j = param_samples[j, "b"]
      r_j = param_samples[j, "r"]
# simulate new noisy observation
      y_j = rnorm(1, mean = pmax, sd = sigma_resid)
      invert_age(y_j, a_j, b_j, r_j)
    })
    age_boot_pi = age_boot_pi[!is.na(age_boot_pi)]
    age_pi_lower[i] = quantile(age_boot_pi, 0.025)
    age_pi_upper[i] = quantile(age_boot_pi, 0.975)
    age_pi_bootstrap[[i]] = age_boot_pi
  }
  
# Append results to data frame
  df$age_point = age_point
  df$age_CI_lower <- age_ci_lower
  df$age_CI_upper <- age_ci_upper
  df$age_PI_lower = age_pi_lower
  df$age_PI_upper = age_pi_upper
  df$age_PI_bootstrap = age_pi_bootstrap
  return(df)
}
```

```{r}
test_genera_dated = predict_age_from_pmax(test_genera_w, "pydamage_pmax", fit_pmax, B = 2000)
head(test_genera_dated)
```


## 5.3 Point estimates and prediction intervals
Point estimates fall somewhere to the line of identity when comparing with rock ages, but further back in time the scatter increases. The prediction intervals are very wide, especially for p_max near the asymptote.

```{r, fig.width=7, fig.height=3.5}
genus_errorbars=ggplot(test_genera_dated[!is.na(test_genera_dated$age_point),], aes(x=median_age_yrs_bf_1950, y=age_point, group=genus))+
  geom_errorbar(aes(ymin=age_PI_lower, ymax=age_PI_upper, color=genus))+
  geom_point(shape=4)+
  theme_bw(base_size=9)+
  geom_abline(intercept = 0, slope = 1, color="black")+
  labs(y='Modelled age', x='Nominal age')+
   scale_color_manual("Genus", values =c( "darkred",   "lightblue3",    '#009E73',    "darkorchid",   "blue4",'#E66101'))+
 facet_wrap(
  ~genus,
  labeller = as_labeller(
    function(x) paste0("italic('", tools::toTitleCase(x), "')"),
    default = label_parsed
  )
)+
theme(legend.position = 'none')
genus_errorbars
```

```{r}
scatter=ggplot(test_genera_dated[!is.na(test_genera_dated$age_point),], aes(x=median_age_yrs_bf_1950, y=age_point, group=genus, color=age_PI_upper-age_PI_lower))+
  geom_point(shape=4)+
  theme_bw(base_size=9)+
  geom_abline(intercept = 0, slope = 1, color="black")+
  labs(y='Modelled Age (Years BP)', x='Nominal Age (Years BP)')+
  theme(legend.position='inside',legend.position.inside = c(0.18, 0.75), legend.key.height = unit(1, "lines"),legend.background = element_rect(fill = "transparent", colour = NA),
legend.box.background = element_rect(fill = "transparent", colour = NA))+
  scale_color_gradientn(colors = c( "black", 'gray40','gray70',"darkred", 'orangered' ),
                        values = scales::rescale(c(5000, 10000, 15000, 20000, 30000), to = c(0,1)),
    breaks  = c(5000, 10000, 15000, 20000, 30000),
    labels  = c('5000', '10000', '15000', '20000', '30000'),
    name = "95% PI width\n(Years)",
    limits  = c(1500, 35000)) 
scatter
 
```

```{r}
modelled = ggplot(test_genera_dated[!is.na(test_genera_dated$age_point),],aes(x = pydamage_pmax, y = age_point)) +
  geom_errorbar(aes(ymin = age_PI_lower,ymax = age_PI_upper,color = "95% prediction interval"),linewidth = 0.3) +
  geom_point(aes(color = "Modelled point age"),shape = 4) +
  scale_color_manual(name = 'Age prediction',values = c("Modelled point age" = "gray40","95% prediction interval" = "lightblue3")) +
  theme_bw(base_size = 9) +
  labs(x = expression(italic(p)[max]),y = "Modelled Age (Years BP)") +
  theme(panel.grid.minor = element_blank(),legend.position = c(0.6, 0.7),legend.justification = c(1, 0),legend.background = element_blank(),legend.margin = margin(4, 6, 4, 6))

modelled
```



```{r}
test_genera_dated[!is.na(test_genera_dated$age_point),] %>% summarise(MAE=mean(abs(median_age_yrs_bf_1950 - age_point)), MRSE=sqrt(mean(abs(median_age_yrs_bf_1950 - age_point)^2)))
```

```{r}
cor.test(test_genera_dated$age_point, test_genera_dated$median_age_yrs_bf_1950, method='spearman')
```


 # 6. Fit logit model for probabilisitic age ranking based on p_max_differences
## 6.1 constructing paired datasets
#### Training data
```{r}
pairs_train = data_w[data_w$median_age_yrs_bf_1950<32000,] %>%
  dplyr::select(sample,median_age_yrs_bf_1950, pydamage_pmax) %>%
  rename_with(~ paste0(.x, "_i")) %>%
  crossing(
    data_w[data_w$median_age_yrs_bf_1950<32000,] %>%
      dplyr::select(sample, median_age_yrs_bf_1950, pydamage_pmax) %>%
      rename_with(~ paste0(.x, "_j"))) %>%
  filter(sample_i != sample_j) %>%
  mutate(
    delta_pmax = pydamage_pmax_j - pydamage_pmax_i,
    y_ij = as.integer(median_age_yrs_bf_1950_j > median_age_yrs_bf_1950_i))
```
```{r, fig.width=3.5}
hist=ggplot(pairs_train, aes(x=delta_pmax, fill=as.factor(y_ij)))+
  geom_histogram(alpha=0.2,position = 'identity', binwidth=0.01)+
  theme_bw(base_size = 9)+
  scale_fill_manual(values=c('gray40','darkred'))+
  scale_x_continuous(expand=c(0,0))+
  scale_y_continuous(expand=c(0,0))+
  labs(x = expression(italic(p)[max*','*j]*' - '*italic(p)[max*','*i] ), y='Count')+
  theme(legend.position='none')+
  annotate('text', x = 0.3, y = 1500, label = expression(Age[j]*'>'*Age[i]), size=3, color='darkred')+
  annotate('text', x = -0.3, y = 1500, label = expression(Age[j]*'<'*Age[i]), size=3, color='gray40')
hist
```

#### Test data
```{r}
pairs_test <- test_genera_w %>%
  dplyr::select(sample, median_age_yrs_bf_1950, pydamage_pmax) %>%
  rename_with(~ paste0(.x, "_i")) %>%
  tidyr::crossing(
    test_genera_w %>%
      dplyr::select(sample, median_age_yrs_bf_1950, pydamage_pmax) %>%
      rename_with(~ paste0(.x, "_j"))) %>%
  filter(sample_i != sample_j) %>%
  mutate(
    y_ij = as.integer(median_age_yrs_bf_1950_j > median_age_yrs_bf_1950_i),
    delta_pmax = pydamage_pmax_j - pydamage_pmax_i)
```
## 6.2 fitting logit model 
```{r}
pmax_logit = glm(
  y_ij ~ delta_pmax,
  family = binomial(),
  data = pairs_train
)
summary(pmax_logit)
```

### 6.2.1 bootstrapping confidence band
```{r}
bootstrap_pairwise_by_genus_sample <- function(
  data,
  time_col = "median_age_yrs_bf_1950",
  damage_col = "pydamage_pmax",
  sample_col = "sample",
  genus_col = "genus",
  R = 100,
  seed = 1963) {
  set.seed(seed)
 #construct pairs from data
  all_pairs <- data %>%
    rename_with(~ paste0(.x, "_i")) %>%
    crossing(
      data %>% rename_with(~ paste0(.x, "_j"))
    ) %>%
    filter(!!sym(paste0(sample_col, "_i")) != !!sym(paste0(sample_col, "_j"))) %>%
    mutate(delta_pmax = !!sym(paste0(damage_col, "_j")) - !!sym(paste0(damage_col, "_i")))
#define range of delta_pmax  
  delta_seq <- seq(
    min(all_pairs$delta_pmax),
    max(all_pairs$delta_pmax),
    length.out = 100)
  # Precompute unique levels and their row indices
  genus_levels = unique(data[[genus_col]])
  sample_levels = unique(data[[sample_col]])
  genus_index_list = lapply(genus_levels, function(g) which(data[[genus_col]]  == g))
  sample_index_list = lapply(sample_levels, function(s) which(data[[sample_col]] == s))
  # Container for predictions
  pred_matrix <- matrix(NA, nrow = R, ncol = length(delta_seq))
  for (r in seq_len(R)) {
    ## 1) Resample genus *clusters* by index, with replacement
    g_ids = sample(seq_along(genus_levels),
                    size = length(genus_levels),
                    replace = TRUE)
    ## 2) Resample sample *clusters* by index, with replacement
    s_ids = sample(seq_along(sample_levels),
                    size = length(sample_levels),
                    replace = TRUE)
    ## 3) Get the row indices for the resampled clusters
    rows_g = unlist(genus_index_list[g_ids])
    rows_s = unlist(sample_index_list[s_ids])
    ## 4) Intersection: rows that belong to both a resampled genus and sample
    rows_keep = intersect(rows_g, rows_s)
    d_resample = data[rows_keep, , drop = FALSE]
    if (nrow(d_resample) < 2) {
      next  # skip iteration if not enough data
    }
    #build pairs and fit glm based on resampled data
    pairs = d_resample %>%
      rename_with(~ paste0(.x, "_i")) %>%
      tidyr::crossing(
        d_resample %>% rename_with(~ paste0(.x, "_j"))
      ) %>%
      dplyr::filter(.data[[paste0(sample_col, "_i")]] != .data[[paste0(sample_col, "_j")]]) %>%
      dplyr::mutate(
        delta_pmax = .data[[paste0(damage_col, "_j")]] -.data[[paste0(damage_col, "_i")]],
        y_ij = as.integer(.data[[paste0(time_col, "_j")]] >.data[[paste0(time_col, "_i")]]))
    if (nrow(pairs) == 0) {
      next}
    mod = glm(y_ij ~ delta_pmax, family = binomial(), data = pairs)
    pred_matrix[r, ] = predict(mod,newdata = data.frame(delta_pmax = delta_seq),type = "response") }
# Compute 95% CI from pred_matrix
ci = apply(pred_matrix, 2, function(x) quantile(x, c(0.025, 0.975), na.rm = TRUE))
ci_df = data.frame(
  delta_pmax = delta_seq,
  lwr = ci[1, ],
  upr = ci[2, ],
  fit = apply(pred_matrix, 2, mean, na.rm = TRUE))
# Return complete results
list(bootstrap_predictions = pred_matrix,ci_df = ci_df)}
```
```{r}
result_two_way = bootstrap_pairwise_by_genus_sample(data = data_w[data_w$median_age_yrs_bf_1950 < 32000, ],R = 500)
```

### 6.2.2 Plot
```{r}
plottable_logit=result_two_way$ci_df %>% mutate(orig_pred = predict(pmax_logit, 
                    newdata = data.frame(delta_pmax = .$delta_pmax), 
                    type = "response"))
```
```{r}
find_cutoff_crossing <- function(df, p) {
  df %>%
    filter(orig_pred >= p | lwr >= p | upr >= p) %>%
    summarise(
      prob = p,
      delta_pmax_fit = min(delta_pmax[orig_pred >= p], na.rm = TRUE),
      delta_pmax_lwr = min(delta_pmax[lwr >= p], na.rm = TRUE),
      delta_pmax_upr = min(delta_pmax[upr >= p], na.rm = TRUE)
    )
}

cut_095 <- find_cutoff_crossing(plottable_logit, 0.95)
cut_075 <- find_cutoff_crossing(plottable_logit, 0.75)

bind_rows(cut_095, cut_075)
```

```{r}
Scurve=ggplot(plottable_logit, aes(x = delta_pmax)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = "95% CI"),  alpha = 0.5) +
  geom_line(aes(x = delta_pmax, y = orig_pred, color="Model fit"), linewidth=0.5)+
  labs(x = expression(italic(p)[max*','*j]*' - '*italic(p)[max*','*i] ), y = expression('P('*Age[j]*'>'*Age[i]*')')) +
  scale_x_continuous(expand=c(0,0))+
  theme_bw(base_size = 9)+
  guides(fill = guide_legend(order = 1),color = guide_legend(order = 1))+
  scale_fill_manual(name='Logistic model',values = c("95% CI" = "lightblue3")) +
  scale_color_manual(name="",values = c("Model fit" = "black"))+
  theme(legend.position = 'none')+
  annotate("rect",xmin = 0.1, xmax = 0.15,ymin = 0.08, ymax = 0.12,fill = "lightblue3",alpha = 0.5) +
  annotate("text",x = 0.18, y = 0.10,label = "95% CI",hjust = 0,size = 3) +
  annotate("segment",x = 0.1, xend = 0.15,y = 0.2, yend = 0.2,linewidth = 0.5) +
  annotate("text",x = 0.18, y = 0.2,label = "Model fit",hjust = 0,size = 3)+
  annotate("text",x = 0.1, y = 0.3,label = "Logistic model",hjust = 0,size = 3.5)
Scurve
```

## 6.3 Application of logit-model
 
```{r}
pairs_test$pred_prob <- predict(
  pmax_logit ,
  newdata = pairs_test,
  type = "response"
)
```
### 6.3.1 ROC
```{r}
roc_obj <- roc(
  response = pairs_test$y_ij,
  predictor = pairs_test$pred_prob
)
```
```{r}
auc_value <- auc(roc_obj)
auc_value
```
```{r}
roc_df <- data.frame(
  specificity = roc_obj$specificities,
  sensitivity = roc_obj$sensitivities
) %>%
  mutate(
    fpr = 1 - specificity
  )

roc=
  ggplot(roc_df, aes(x = fpr, y = sensitivity)) +
  geom_line(color = "black", size = 1) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "grey50") +
  scale_x_continuous(limits = c(0, 1), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate") +
  theme_bw(base_size = 9)+
  annotate( "text",x = 0.6, y = 0.2,label = paste0("AUC = ", round(auc_value, 3)),size = 3)+
  coord_fixed()
roc
```

### 6.3.2 Classification threshold
```{r}
coords(roc_obj, "best", ret = c("threshold","sensitivity", "specificity"), best.method="youden")
```

```{r}
qlogis(c(0.5)) / coef(pmax_logit)['delta_pmax']
```
0.5 is the most suited threshold for class separation, this corresponds to a delta_pmax of 0 (as expected for logit model)

```{r}
pairs_test$pred_class <- pairs_test$pred_prob > 0.5
```

```{r}
table(
  truth = pairs_test$y_ij,
  pred  = pairs_test$pred_class
)

mean(pairs_test$y_ij == pairs_test$pred_class)
```

```{r}
mean((pairs_test$delta_pmax > 0) == pairs_test$y_ij)
```

### 6.3.3 Binned data approach

```{r}
pairs_test_plot <- pairs_test %>%
  mutate( correct = as.integer((delta_pmax > 0) == (median_age_yrs_bf_1950_j>median_age_yrs_bf_1950_i)), abs_dp = abs(delta_pmax),
      dp_bin = cut_width(abs_dp, width = 0.025, boundary = 0),include.lowest = TRUE)%>% 
 mutate(correct = factor(correct, levels = c(0, 1),
                     labels = c("Incorrect", "Correct"))) %>% 
  group_by(dp_bin, correct) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(dp_bin) %>%
  mutate(prop = n / sum(n))
```
```{r}
bars=ggplot(pairs_test_plot, aes(x = dp_bin, y = prop, fill = correct)) +
  geom_col(width = 0.9, alpha=0.5) +
  labs(x = expression("|"*Delta*"("*italic(p)[max]*")|"),y = "Probability of correct ranking",fill = NULL) +
  scale_fill_manual(name='Classification',values = c("Incorrect" = "darkred", "Correct" = "grey40")) +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position=c(0.5,0.91), legend.box = 'horizontal',legend.direction = "horizontal",legend.key.width = unit(0.4, "cm"),legend.key.height = unit(0.35, "cm"),legend.background = element_rect(fill = NA, color = NA))+
  guides(fill = guide_legend(ncol = 2))+
  scale_y_continuous(limits = c(0, 1.2),breaks = c(0.25, 0.5, 0.75, 1)
)
bars
```

# 7. combined plots
## 7.1 Figure 3
```{r, fig.width=7.086614, fig.height=3.5}
cor2 <- cor +
  theme(
    axis.title.y = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )

assembly2 <- assembly +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    plot.margin = margin(0, 0, 0, 0)
  )

cor2 + assembly2 +
  plot_layout(widths = c(3.5, 1)) +
  plot_annotation(tag_levels = "A")
```
```{r}
ggsave('../06_figures/Fig3/fig3.pdf', width=18, height=8.5, units='cm',create.dir = TRUE)
```

## 7.2 Figure 5
```{r, fig.width=7, fig.height=5.5}
((p_mode | p_pgeom) / (p_lambda | p_cor)) +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "right",
    plot.margin = grid::unit(c(0, 0, 0, 0), "pt")
  )
```

```{r}
ggsave('../06_figures/Fig5/fig5.pdf', width=18, height=14, units='cm',create.dir = TRUE)
```

## 7.3 Figure 6
```{r, fig.width=7, fig.height=3}
(p_deltas | p_pmax) + plot_annotation(tag_levels = "A")+ plot_layout(guides = "collect") &  theme(legend.position = "right")
```

```{r}
ggsave('../06_figures/Fig6/fig6.pdf', width=18, height=7.5, units='cm',create.dir = TRUE)
```

## 7.4 Figure 7
```{r,fig.width=7, fig.height=7}
(overview/(modelled|scatter))+plot_annotation(tag_levels = "A")
```
```{r}
ggsave('../06_figures/Fig7/fig7.pdf', width=18, height=18, units='cm',create.dir = TRUE)
```

## 7.5 Figure 8
```{r, fig.width=7, fig.height=5}
(hist | Scurve) / (roc | bars) + plot_annotation(tag_levels = "A")
```
```{r}
ggsave('../06_figures/Fig8/fig8.pdf', width=18, height=12.5, units='cm',create.dir = TRUE)
```
## 7.6 Supplementary Figure 3
```{r}
ggsave('../06_figures/Supplementary_Fig3.png',comparison, width=18, height=7.5, units='cm', dpi='print',create.dir = TRUE)
```
## 7.7 Supplementary Figure 4
```{r}
ggsave('../06_figures/Supplementary_Fig4.png',scatter_cor, width=18, height=22.5, units='cm', dpi='print',create.dir = TRUE)
```
## 7.8 Supplementary Figure 7
```{r}
ggsave('../06_figures/Supplementary_Fig7.png',genus_errorbars, width=18, height=10, units='cm', dpi='print',create.dir = TRUE)
```

## 7.9 Supplementary Figure 8 
```{r, fig.width= 5, fig.height= 9}
(delta_s_res+
  theme(axis.title.x = element_blank()))/ 
(pmax_res+
   theme(axis.title.x = element_blank()))/ 
(mode_res+
   theme(axis.title.x = element_blank()))/ 
(p_geom_res+
   theme(axis.title.x = element_blank()))/ 
lambda_res + plot_annotation(tag_levels = "A")+ plot_layout(guides = "collect")&theme(legend.position = "bottom",  plot.margin = margin(0, 0, 0,0)) 

ggsave('../06_figures/Supplementary_Fig8.png', width=18, height=22.5, units='cm', dpi='print',create.dir = TRUE)
 
```
## 7.10 Supplementary Figure 9
```{r}
ggsave('../06_figures/Supplementary_Fig9.png',gc_plot, width=18, height=7.5, units='cm', dpi='print',create.dir = TRUE)
```
## 7.11 Supplementary Figure 10
```{r, fig.width=7, fig.height=8}
wrap_plots(results_list_gc[[2]] , results_list_gc[[4]],results_list_gc[[3]], results_list_gc[[5]],results_list_gc[[1]],patchwork::guide_area(),ncol=2 )+plot_annotation(tag_levels = "A")+ plot_layout(guides = "collect")&theme(legend.position = 'bottom', legend.direction = 'vertical', legend.margin = margin(0,0,0,0),plot.margin = margin(2, 2, 2, 2),legend.box.margin = margin(0, 0, 0, 0),legend.spacing.y = unit(1, "pt"),legend.title = element_text(margin = margin(r = 10)))&guides(color = guide_legend(ncol = 1, order = 1), shape = guide_legend(order = 2, ncol = 1))

ggsave('../06_figures/Supplementary_Fig10.png', width=18, height=20, units='cm', dpi='print',create.dir = TRUE)
```
# 8 Source data files

```{r}
plottable_cor=cor_data[,c('genus','tool_param','estimate','p_value','p_adj')]
write.csv(plottable_cor, '../07_output_tables/Fig3A_source.txt', quote=F, row.names = FALSE)
```

```{r}
write.csv(data_w[,c('sample','genus','pydamage_len')], '../07_output_tables/Fig3B_source.txt', quote=F, row.names = FALSE)
```

```{r}
write.csv(plottable_mode, file = "../07_output_tables/Fig5A_source.txt", row.names = FALSE)
```

```{r}
write.csv(plottable_pgeom, file = "../07_output_tables/Fig5B_source.txt", row.names = FALSE)
```

```{r}
write.csv(plottable_overhangs, file = "../07_output_tables/Fig5C_source.txt", row.names = FALSE)
```

```{r}
write.csv(data_w[,c('mapdamage_1_over_lambda','compute_length_geom_p')], file = "../07_output_tables/Fig5D_source.txt", row.names = FALSE)
```

```{r}
write.csv(plottable_delta_s, file = "../07_output_tables/Fig6A_source.txt", row.names = FALSE)
```

```{r}
write.csv(plottable_pmax, file = "../07_output_tables/Fig6B_source.txt", row.names = FALSE)
```

```{r}
write.csv(plottable_overview, '../07_output_tables/Fig7A_source.txt', quote=F, row.names = FALSE)
```

```{r}
plottable_ages_errorbars=test_genera_dated[!is.na(test_genera_dated$age_point),c('sample','genus','pydamage_pmax','age_point','age_PI_lower','age_PI_upper')]
write.csv(plottable_ages_errorbars, '../07_output_tables/Fig7B_source.txt', quote=F, row.names = FALSE)
```

```{r}
plottable_scatter=test_genera_dated[!is.na(test_genera_dated$age_point),c('sample','genus','median_age_yrs_bf_1950', 'age_point', 'age_PI_upper','age_PI_lower')] %>%mutate(PI_width=age_PI_upper-age_PI_lower)
write.csv(plottable_scatter, '../07_output_tables/Fig7C_source.txt', quote=F, row.names = FALSE)
```

```{r}
plottable_hist= pairs_train %>%dplyr::select(delta_pmax, y_ij) %>%
  mutate(age_order = ifelse(y_ij == 1,"Age[j] > Age[i]","Age[j] < Age[i]"))
write.csv(plottable_hist, '../07_output_tables/Fig8A_source.txt', quote=F, row.names = FALSE)
```

```{r}
write.csv(plottable_logit[,c('delta_pmax', 'upr','lwr','orig_pred')], '../07_output_tables/Fig8B_source.txt', quote=F, row.names = FALSE)
```

```{r}
write.csv(roc_df[,c('fpr','sensitivity')], '../07_output_tables/Fig8C_source.txt', quote=F, row.names = FALSE)
```

```{r}
write.csv(pairs_test_plot, '../07_output_tables/Fig8D_source.txt', quote=F, row.names = FALSE)
```