---
title: "TPM based community structure of stalagmite TM-17"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
In this notebook I run and visualize a PCA on the taxonomic composition of the TM-17 aDNA assemblies and extraction 
# 1. Preparations
```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
getwd()  # check it worked
```

## 1.1 Loading packages
```{r}
library(tidyverse)
packageVersion('tidyverse')
library(vegan)
packageVersion('vegan')
library(ggrepel)
packageVersion('ggrepel')
library(patchwork)
packageVersion('patchwork')
```
## 1.2 Loading data (prepared in python)
```{r}
data=read.csv("../01_concatenate_mmseqs_output/2025-10-07_contig_tax_tpm.csv", sep=';')
```


# 2 PCA of community composition (based on taxonomic identifier)
## 2.1 Taxon abundance table
```{r}
taxa_wide= data %>% group_by(sample, taxon) %>% summarize(summed_TPM=sum(tpm))%>% mutate(across(where(is.character), ~ifelse(. == "", "unclassified", .)))%>% pivot_wider(names_from = taxon, values_from = summed_TPM, values_fill = 0,values_fn = sum) %>% mutate(across(where(is.numeric), ~replace_na(., 0))) %>% ungroup() %>% as.data.frame()

rownames(taxa_wide)=taxa_wide[,1]
taxa_wide=taxa_wide[,-1]
```

## 2.2 PCA on Hellinger-transformed abundances
```{r}
tax.pca <- rda(decostand(taxa_wide, method = "hell"))
tax.pca
```
```{r, fig.height=4, fig.width=4}
tax_var=sprintf("%.1f %%", tax.pca$CA$eig/tax.pca$tot.chi*100)
tax_var

df <- as.data.frame(scores(tax.pca)$sites) %>% rownames_to_column("Sample") %>%
  mutate(sample_type = substr(Sample, 1, 3))
```

## 2.3 Plot PCA Biplot
```{r}
pca_plot=ggplot(df, aes(x = PC1, y = PC2, color = sample_type)) +
  geom_point(alpha=0.5) +
  scale_color_manual("Sample Type",labels = c('TDM'="stalagmite",'EXB'= "extraction blank"), values=c('lightblue3','darkred'))+
  stat_ellipse(type = "t", level = 0.95) +  
  xlab(bquote("PC1 ("*.(tax_var[1])*" variance)")) +
  ylab(bquote("PC2 ("*.(tax_var[2])*" variance)")) +
  theme_bw(base_size = 9)+
  theme(legend.position = 'bottom')+
   geom_path(aes(group = sample_type),color='gray40',alpha=0.5, arrow = arrow(length = unit(0.2, "cm"), type = "closed"))+
  geom_label_repel(data=df[df$Sample %in% c('TDM001','TDM061', 'TDM056', 'TDM024', 'TDM037', 'TDM035'),],aes(label=Sample),size = 2.5,
    segment.color = "darkred",
    segment.size = 0.5,
    min.segment.length = 0,
    direction='both', box.padding=0.75, show.legend = FALSE, max.overlaps = 5)
pca_plot
```

# 3. Volcano plot
The purpose of this analysis is to identify stalagmite enriched taxa in comparison to the extraction blank. I'm focussing on the subset of contigs annotated at genus level. q-values shown on the y-axis are obtained from a permutation test.

## 3.1 Genus abundance table 
```{r}
genus_wide= data %>% group_by(sample, genus) %>% summarize(summed_TPM=sum(tpm))%>% mutate(across(where(is.character), ~ifelse(. == "", "unclassified", .)))%>% pivot_wider(names_from = genus, values_from = summed_TPM, values_fill = 0,values_fn = sum) %>% mutate(across(where(is.numeric), ~replace_na(., 0))) %>% ungroup() %>% as.data.frame()

rownames(genus_wide)=genus_wide[,1]
genus_wide=genus_wide[,-1]
```
## 3.2 Calculate prevalence of genera
```{r}
compute_prevalence= function(abund_df, control_prefix = "EXB", sample_prefix = "TDM", epsilon = 1e-6) {
  # Identify which rows are controls and which are samples
  control_rows = grepl(paste0("^", control_prefix), rownames(abund_df))
  sample_rows  = grepl(paste0("^", sample_prefix), rownames(abund_df))
  # Compute prevalence (fraction of samples with non-zero abundance)
  prev_control = colSums(abund_df[control_rows, , drop = FALSE] > 0, na.rm = TRUE)
  prev_sample  = colSums(abund_df[sample_rows, , drop = FALSE] > 0, na.rm = TRUE)
  # Combine results into data frame
  prevalence_df = data.frame(
    Taxon = names(prev_sample),
    prev_control = prev_control,
    prev_sample  = prev_sample,
    stringsAsFactors = FALSE
  )
  return(prevalence_df)
}

```
```{r}
genus_prevalence_df=compute_prevalence(genus_wide)
```

## 3.3 CLR transformation
```{r}
clr_transform = function(df, pseudocount = 1e-6) {
  df = df + pseudocount
  log_df = log(df)
  gm_mean = rowMeans(log_df)
  clr_df = sweep(log_df, 1, gm_mean, "-")
  return(clr_df)}
```
```{r}
clr_df = clr_transform(genus_wide) %>%
  as.data.frame() %>%
  mutate(lib_type = substr(rownames(.), 1, 3))
```

## 3.4 Permutation test
```{r}
perm_test_all_genera = function(data, lib_type_col = "lib_type",
                                 control_label = "EXB",
                                 sample_label = "TDM",
                                 n_perm = 10000) {
  # Ensure lib_type column exists
  if (!(lib_type_col %in% names(data))) {
    stop("Column '", lib_type_col, "' not found in data.")
  }

  # Identify which columns are genera (numeric only)
  genus_cols = names(data)[sapply(data, is.numeric)]
  genus_cols = setdiff(genus_cols, lib_type_col)

  # Split data into controls and samples
  controls_idx = data[[lib_type_col]] == control_label
  samples_idx  = data[[lib_type_col]] == sample_label

  results = data.frame(
    genus = genus_cols,
    obs_diff = NA_real_,
    pval = NA_real_,
    stringsAsFactors = FALSE
  )

  # Iterate through each genus
  for (g in genus_cols) {
    controls <- data[controls_idx, g]
    samples  <- data[samples_idx, g]

    # skip genera with NA or too few values
    if (anyNA(controls) || anyNA(samples) || length(controls) < 2) next

    obs_diff = mean(samples) - mean(controls)
    combined = c(samples, controls)
    n_samp = length(samples)

    perm_diffs = numeric(n_perm)
    for (i in seq_len(n_perm)) {
      shuffled = sample(combined)
      perm_diffs[i] = mean(shuffled[1:n_samp]) - mean(shuffled[(n_samp + 1):length(combined)])
    }

    pval = (sum(abs(perm_diffs) >= abs(obs_diff)) + 1) / (n_perm + 1)

    results[results$genus == g, c("obs_diff", "pval")] <- c(obs_diff, pval)
  }

  # Add FDR-corrected q-values
  results$qval = p.adjust(results$pval, method = "fdr")

  return(results)
}

```
For reproducibility and time reasons I saved the results of the permutation test for later use.
```{r}
set.seed(6397)
#results_high_perm =  perm_test_all_genera(clr_df,n_perm = 100000)
```
```{r}
#write.csv(results_high_perm, file = "results_high_perm.csv", row.names = FALSE)
```
```{r}
results_high_perm=read.csv("./results_high_perm.csv")
```

## 3.5 Plot Volcano plot
```{r,fig.width=4, fig.height=4}
volcano=results_high_perm %>% merge(genus_prevalence_df, by.x='genus', by.y='Taxon') %>% 
  mutate(
    prev_bin = cut(
      prev_sample,
      breaks = c(-Inf, 6,11 , 31, 50, 55, Inf),
      labels = c('0-5',"6-10","11-30", "31-49", "50-54", "55-60"),
      right = FALSE
    )) %>% 
ggplot(aes(x = obs_diff, y = -log(qval))) +
  geom_point(aes(color=prev_bin),alpha=0.5)+
  theme_bw(base_size = 9)+
  scale_color_manual(
    name = "Prevalence\n(# stalagmite samples)",
    values = c(
      "0-5"=  "slategray4",
      "6-10"   = "lightblue3",
      "11-30" =  "lightblue1",
      "31-49" = "#DBC8C8",
      "50-54" = "firebrick2",
      "55-60" = "darkred"
    ))+
  labs(
    x = expression(bar(CLR)[stalagmite]~-~bar(CLR)[blank]),
    y = "-log(q-value)"
  )+
  geom_label_repel( data = results_high_perm[results_high_perm$obs_diff > 19 &
                                 results_high_perm$qval<0.01, ],
    aes(label = genus), size=2.5, force=50, max.overlaps = 3,point.padding=0,direction='y', nudge_y=0.5)+
  ylim(c(0,10))+
  xlim(c(-25,25))+
  theme(legend.position = 'bottom')
volcano
```
```{r}
t=results_high_perm %>% merge(genus_prevalence_df, by.x='genus', by.y='Taxon') %>% 
  mutate(
    prev_bin = cut(
      prev_sample,
      breaks = c(-Inf, 6,11 , 31, 50, 55, Inf),
      labels = c('0-5',"6-10","11-30", "31–49", "50–54", "55–60"),
      right = FALSE
    )) 
t
```

# 4. Combined Plot (Figure 2)
```{r, fig.width=7}
(pca_plot|volcano) + plot_annotation(tag_levels = "A")
#ggsave('../../../06_figures/Fig2/Fig2.pdf')
```

# 5. Taxonomic overview table for Supplementary

```{r}
S1= data %>% group_by(taxID,taxon, rank,kingdom,phylum,class,order,family,genus,species,sample) %>% summarize(summed_TPM=sum(tpm))%>% mutate(across(where(is.character), ~ifelse(. == "", "unclassified", .)))%>% pivot_wider(names_from = sample, values_from = summed_TPM, values_fill = 0,values_fn = sum) %>% mutate(across(where(is.numeric), ~replace_na(., 0))) %>% ungroup() %>% as.data.frame()
```
```{r}
write.csv(S1, file = "../../../07_overview_tables/S1_tax_overview.csv", row.names = FALSE)
```


